# Download the MICS Data {#c01-download}

This chapter shows how to download official MICS microdata from UNICEF
and unpack it into a clean folder tree. MICS datasets are released as
**SPSS** (`.sav`) files. They already carry variable and value labels,
so you can open them directly in R or Stata without separate label codebooks or
scripts, or a format conversion.

## Download from UNICEF

Official MICS microdata are published on the UNICEF MICS surveys page.

1.  Open [https://mics.unicef.org/surveys](https://mics.unicef.org/surveys).
2.  Log in if the site asks you to (dataset downloads often require an
    account).
3.  Set the filters to match the surveys you want, for example, round
    (**MICS6**), region or countries, data type (**MICS**), and status
    (**Completed**).
4.  Click **Download MICS datasets**. The site packages every survey that
    matches your filters into one bulk zip (often named
    `MICS_Datasets.zip`).

Save the zip somewhere easy to find, such as your Downloads folder.

## Messy nested survey archives

UNICEF’s bulk zip packs each survey in its own folder, and each folder
usually holds another zip of SPSS files. Unpacking that by hand across
many countries wastes time and invites path errors.

What you want instead is a flat tree: one folder per survey, with the
`.sav` files and readme sitting directly inside it.

``` text
data/MICS_Datasets/
├── BEN_2021_MICS6_v01_M/
│   ├── bh.sav
│   ├── ch.sav
│   ├── fs.sav      # children aged 5–17 (foundation skills / FS module)
│   ├── hh.sav
│   ├── hl.sav
│   ├── mn.sav
│   ├── tn.sav
│   ├── wm.sav
│   └── Readme_....rtf   # or "Read me_....txt" — name varies by survey
├── GHA_2017_MICS6_v01_M/
│   └── ...
└── ...
```

Each country-year folder keeps **all** SPSS modules from that release,
plus the readme. The prepare script below produces this layout from
whatever surveys sit inside your bulk zip.

## Prepare the folders with `prepare-mics-fs.R`

AFLEARN’s
[`prepare-mics-fs.R`](https://github.com/datafirst-aflearn/mics-guide/blob/main/R/prepare-mics-fs.R)
turns UNICEF’s messy nested survey archives into that clean folder tree.
The script finds `MICS_Datasets.zip` in your working directory, project,
or Downloads; extracts the `.sav` files and readmes; strips nested
archives and wrapper folders; and writes `data/MICS_Datasets/{SURVEY}/`.
It needs only base R, and it runs beside the zip in any folder—Downloads,
Desktop, a USB drive, or a project directory.

Download the script, copy it next to `MICS_Datasets.zip`, set that folder
as the working directory in R, and run:

```{r eval=FALSE}
setwd("C:/path/to/folder/with/zip/and/script")
source("prepare-mics-fs.R")
```

The script writes path-ready survey files to `data/MICS_Datasets/` under
that folder. The original bulk zip is **kept** by default. To delete it
after a successful run:

```{r eval=FALSE}
prepare_mics_data(remove_zip = TRUE)
```

If the zip lives elsewhere, source the script once to load the functions,
then call:

```{r eval=FALSE}
prepare_mics_data(zip = "C:/path/to/MICS_Datasets.zip")
```


## Quality checks

Before you analyse, check that:

1.  `data/MICS_Datasets/` exists under the folder you used as the output
    root (for this guide, the project folder).
2.  At least one survey folder is present (for example
    `data/MICS_Datasets/BEN_2021_MICS6_v01_M/`).
3.  That folder contains `.sav` files (including `fs.sav` when the survey
    fielded the children 5–17 module) and usually a readme.
4.  You can open a `.sav` file in R (`haven::read_sav()`) or Stata
    (`import spss`).

```{r eval=FALSE}
# R
dir.exists("data/MICS_Datasets")     # 1
list.files("data/MICS_Datasets")     # 2
file.exists("data/MICS_Datasets/BEN_2021_MICS6_v01_M/fs.sav")  # 3
```

```{stata eval=FALSE}
* Stata
confirmdir "data/MICS_Datasets" # 1 and # 2
confirm file "data/MICS_Datasets/BEN_2021_MICS6_v01_M/fs.sav"  # 3
```

## Next steps

Once `data/MICS_Datasets/` is in place—and especially once the supported
surveys have `fs.sav`—continue with
[Harmonising MICS6 reading outcomes](#c05-harmonize). For survey design and
file structure, see [Understanding MICS Data](#ch-whats-in-data).
