# Download the MICS Data {#c01-download}

Before you can follow the R or Stata examples in this guide, you need
the official MICS microdata on your computer in the right place.

This chapter walks you through:

1.  Where to download the official data
2.  Which files you need for this guide
3.  How to organise your folders with the prepare script
4.  A short checklist before you start analysing

MICS datasets are released as **SPSS** (`.sav`) files. They already
include variable and value labels, so you do not need separate label
scripts, and you do not need to convert the files to Stata format before
starting.

## What you need for this guide

The analysis chapters assume you are working from this guide’s project
folder and that prepared survey folders exist under
`data/MICS_Datasets/`. After you run the prepare script (Step 3), a
typical layout looks like this:

``` text
data/MICS_Datasets/
├── BEN_2021_MICS6_v01_M/
│   ├── bh.sav
│   ├── ch.sav
│   ├── fs.sav      # children aged 5–17 (foundation skills / FS module)
│   ├── hh.sav
│   ├── hl.sav
│   ├── mn.sav
│   ├── tn.sav
│   ├── wm.sav
│   └── Readme_....rtf   # or "Read me_....txt" — name varies by survey
├── GHA_2017_MICS6_v01_M/
│   └── ...
└── ...
```

Each survey folder keeps **all** SPSS modules from that release, plus the
readme. Module sets differ slightly by country (for example some surveys
include `fg.sav`). The **FS** file (`fs.sav`) holds the questionnaire for
children aged 5–17, which is the unit of analysis used in this guide’s
worked examples.

Because the `.sav` files are already labelled, you can open them directly
in R with `haven::read_sav()` or in Stata with `import spss`.

## Download the data from UNICEF MICS

Official MICS microdata are published on the UNICEF MICS surveys page.

1.  Open [https://mics.unicef.org/surveys](https://mics.unicef.org/surveys).
2.  Log in if the site asks you to (dataset downloads often require an
    account).
3.  Set the filters to match the surveys you want. For this guide’s
    Africa MICS6 examples, use:
    -   Round: **MICS6**
    -   Region / countries: **African countries** (or the countries you
        need)
    -   Data type: **MICS**
    -   Status: **Completed**
4.  Click **Download MICS datasets**. The site packages all surveys that
    match your filters into one bulk zip (often named
    `MICS_Datasets.zip`).

You can use other filters (for example MICS7 only, or a single country).
The prepare script in Step 3 is **not** limited to one round or to a
fixed list of countries: it organises whatever surveys are inside the
bulk zip, as long as UNICEF uses the usual folder-and-nested-zip layout.

Save the downloaded zip somewhere easy to find (for example your
Downloads folder, or this guide’s project folder).

## Set up your project folders

Work from this guide’s project root (the folder that contains `index.Rmd`
and the chapter `.Rmd` files). Create a `data/` folder there if it does
not already exist. The prepare script lives in `R/`:

``` text
mics-guide/
├── data/                         # created if needed; prepared surveys go here
├── R/
│   └── prepare-mics-fs.R         # already in the guide
├── 01-download.Rmd
├── 02-whats-in-the-data.Rmd
├── 03-r.Rmd
└── 04-stata.Rmd
```

Keeping prepared files under `mics-guide/data/MICS_Datasets/` means the
analysis chapters can use short relative paths. If you prefer another
location, that is fine, but then you must change the file paths in the R
and Stata chapters to match.

## Run the prepare script

The bulk zip contains one folder per survey, and each of those folders
contains another zip with the SPSS files. Unpacking every nested archive
by hand is tedious. This guide includes an R script that does it for you:
it finds `MICS_Datasets.zip`, extracts every `.sav` file and the readme
for each survey, and writes a clean folder tree under
`data/MICS_Datasets/`.

### Option A: Run from the guide project

1.  Put `MICS_Datasets.zip` in your **mics-guide** project root, or in `data/`.
2.  In R, set the working directory to the **mics-guide** project root.
3.  Run:

```{r eval=FALSE}
source("R/prepare-mics-fs.R")
```

The script searches for the zip in the working directory, the project
`data/` folder and the project root. If you
need to point to a zip in an unusual location, source the script once
(so the functions are loaded), then call:

```{r eval=FALSE}
prepare_mics_data(zip = "C:/Users/you/path/MICS_Datasets.zip")
```

### Option B: Script and zip in the same folder

You can copy `R/prepare-mics-fs.R` next to `MICS_Datasets.zip` in any
folder on your computer (Downloads, Desktop, a USB drive, and so on):

```{r eval=FALSE}
setwd("C:/path/to/folder/with/zip/and/script")
source("prepare-mics-fs.R")
```

Organised files are written to `data/MICS_Datasets/` under that folder
(or under the mics-guide project root if the script can find the
project).

The original bulk zip is **kept** by default. To delete it after a
successful run, call `prepare_mics_data(remove_zip = TRUE)`.

## Step 4: Confirm everything is ready

Before moving on to the R or Stata chapters, check that:

1.  `data/MICS_Datasets/` exists in the guide project folder.
2.  At least one survey folder is present (for example
    `data/MICS_Datasets/BEN_2021_MICS6_v01_M/`).
3.  That folder contains `.sav` files (including `fs.sav` when the survey
    fielded the children 5–17 module) and usually a readme.
4.  You can open a `.sav` file in R (`haven::read_sav()`) or Stata
    (`import spss`).
5.  Your R or Stata working directory is the guide project root when you
    knit or run the examples.

Quick checks:

```{r eval=FALSE}
# R
dir.exists("data/MICS_Datasets")
list.files("data/MICS_Datasets")
file.exists("data/MICS_Datasets/BEN_2021_MICS6_v01_M/fs.sav")
```

```{stata eval=FALSE}
* Stata
confirmdir "data/MICS_Datasets"
confirm file "data/MICS_Datasets/BEN_2021_MICS6_v01_M/fs.sav"
```

If these checks succeed (using a survey folder name that exists on your
machine), you are ready for the analysis chapters.

## Next steps

Once `data/MICS_Datasets/` is in place:

-   Follow [Understanding MICS Data](#ch-whats-in-data) for survey design
    and file structure.
-   Follow [R Tutorial](#c03-r) if you work in R.
-   Follow [Stata Tutorial](#c04-stata) if you work in Stata.
