## Getting UNICEF MICS files {#c01-unicef}

UNICEF releases MICS microdata as SPSS (`.sav`) files. The files already carry variable and value labels, so you can open them in R or Stata without a separate codebook.

### Download the country files

1.  Open <https://mics.unicef.org/surveys>.
2.  Log in if the site asks you to. Dataset downloads usually need an account.
3.  Filter to the surveys you want: round **MICS6**, then region or countries, data type **MICS**, and status **Completed**.
4.  Click **Download MICS datasets**. UNICEF packs every matching survey into one zip, usually named `MICS_Datasets.zip`.

Make a project folder, for example `MICS project`, and save `MICS_Datasets.zip` in it.

### Unpack with AFLEARN's scripts

The bulk zip nests each survey in further archives and wrapper folders. Unpacking many countries by hand wastes time and invites path errors.

``` text
MICS_Datasets.zip
└── survey folder
    └── another archive
        └── wrapper folder
            ├── fs.sav
            ├── hh.sav
            └── ...
```

Download [Script.zip](https://github.com/datafirst-aflearn/mics-guide/raw/main/Script.zip) and unzip it into the same project folder, so `Script/` sits beside `MICS_Datasets.zip`:

``` text
MICS project/
├── MICS_Datasets.zip
└── Script/
```

In R, set the project folder as the working directory and source `run-it.R`:

```{r eval=FALSE}
setwd("C:/path/to/MICS project")
source("Script/run-it.R")
```

The working directory must be the folder that holds `Script/`. If it is not, the script stops before creating anything.

That one run unpacks the surveys and writes the AFLEARN analysis file. Later chapters explain those files. You do not need to source the other scripts on the first run.

After a successful run:

``` text
MICS project/
├── MICS_Datasets.zip
├── Script/
└── Data/
    ├── UNICEF/
    │   ├── BEN_2021_MICS6_v01_M/
    │   │   ├── fs.sav
    │   │   ├── hh.sav
    │   │   ├── hl.sav
    │   │   └── ...
    │   └── ...
    └── AFLEARN Harmonised Data/
        ├── mics6-crosswalk.xlsx
        └── mics6-fs-harmonised-v2.0.dta
```

Each survey then has a predictable path such as `Data/UNICEF/BEN_2021_MICS6_v01_M/fs.sav`.

If the UNICEF zip is not in the project folder, pass its path:

```{r eval=FALSE}
source("Script/prepare-mics-fs.R")
prepare_mics_data(zip = "C:/path/to/MICS_Datasets.zip")
```

Then source `Script/run-it.R` as usual, or source the remaining scripts from `Script/` yourself.

### Check the result

```{r eval=FALSE}
dir.exists("Data/UNICEF")
list.files("Data/UNICEF")
file.exists("Data/UNICEF/BEN_2021_MICS6_v01_M/fs.sav")
```

Open one SPSS file:

```{r eval=FALSE}
library(haven)
fs <- read_sav("Data/UNICEF/BEN_2021_MICS6_v01_M/fs.sav")
dim(fs)
```

::: alternativebox-header
**Keep the original download**
:::

::: alternativebox
Keep the original UNICEF zip. The clean folders are for analysis; the zip is the unchanged source.
:::

[Understanding MICS6 data](#ch-whats-in-data) explains the survey files and the FLS module.
