
# Analysing MICS6 data in R {#c03-r}

UNICEF’s Multiple Indicator Cluster Surveys (MICS) collect comparable
information on children across many African countries. Official country files
use different layouts and codes, so they do not pool cleanly on their own. This
guide’s harmonised file puts those surveys on one child-level footing: shared
names, a common education-level map, numeracy items, and reading outcomes.

This chapter asks what that African file can show about schooling and
foundational skills. The ladder is fixed: explore the sample, check over-age
enrolment across countries, then dig into Ghana for private schooling,
numeracy, and a gender interaction. Download and harmonisation steps live in
[Getting MICS6 data](#c01-download) and
[Harmonising MICS6 data](#c03-harmonise).

## Packages

Install once if needed, then load at the start of every session:

```{r install-pkgs, eval=FALSE}
install.packages(c("tidyverse", "haven", "srvyr", "survey", "showtext", "sysfonts"))
```

```{r setup-r, include=FALSE}
mics_file <- "Data/AFLEARN Harmonised Data/mics6-fs-harmonised-v2.0.dta"
has_mics <- file.exists(mics_file)
has_r_pkgs <- requireNamespace("tidyverse", quietly = TRUE) &&
  requireNamespace("haven", quietly = TRUE) &&
  requireNamespace("srvyr", quietly = TRUE) &&
  requireNamespace("survey", quietly = TRUE)
run_r_lab <- has_mics && has_r_pkgs
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 9,
  fig.height = 5,
  fig.align = "center",
  eval = run_r_lab
)
```

```{r packages-r}
library(tidyverse)
library(haven)
library(srvyr)
library(survey)   # svyglm on srvyr designs

source("R/aflearn_theme.R")
```


## Load the harmonised file

If you do not yet have the file, start from [Getting UNICEF MICS files](#c01-unicef). After `read_dta()`, you hold one row per selected child in `fs`: schooling background, sex, sample weights, numeracy items, and reading outcomes where the assessment was completed.

```{r load-r}
dat <- read_dta("Data/AFLEARN Harmonised Data/mics6-fs-harmonised-v2.0.dta")

num_items <- intersect(
  c(
    paste0("number_id_", 1:6),
    paste0("number_compare_", 1:5),
    paste0("number_add_", 1:5),
    paste0("number_pattern_", 1:5)
  ),
  names(dat)
)
items <- dat[num_items]
dat$numeracy_score <- if_else(
  length(num_items) == 0L | rowSums(!is.na(items)) == 0,
  NA_real_,
  rowSums(items == 1, na.rm = TRUE)
)

dat <- dat %>%
  mutate(
    survey = paste(country_iso3, year),
    female = if_else(sex == 2L, 1L, if_else(sex == 1L, 0L, NA_integer_)),
    private = case_when(
      country_iso3 == "COD" ~ NA_integer_,
      school_type == 3 ~ 1L,
      school_type %in% c(1, 2, 4, 6) ~ 0L,
      TRUE ~ NA_integer_
    )
  )

glimpse(dat %>% select(
  country_iso3, year, HH1, HH2, LN, age, female, private, enrolled,
  current_level_h, current_grade, numeracy_score, reading_skills, fsweight
))
```

`numeracy_score` counts correct answers across the 21 foundational numeracy
items (0–21). `reading_skills` is 1 when a child met the foundational reading
threshold among those assessed. `private` is 1 for private schools and 0 for
public, religious, community, or other under the usual MICS ownership codes
(COD is left missing).

## Explore and unique child ID

Before any model, confirm what each row is and that children do not duplicate.
A broken key silently double-counts households when you merge or tabulate.

```{r explore-id}
nrow(dat)
ncol(dat)
count(dat, country_iso3, year)

# Candidate keys
key_vars <- c("country_iso3", "year", "HH1", "HH2", "LN")
n_distinct(dat[key_vars]) == nrow(dat)

# Same idea with the compact ID copies
key_vars2 <- c("country_iso3", "year", "cluster", "hhno", "linech")
n_distinct(dat[key_vars2]) == nrow(dat)
```

The country–year table is the map of this guide’s Africa coverage: nineteen
surveys from Benin through Zimbabwe (Tunisia appears twice). Both key
combinations return `TRUE` — one row per child. A single `childid` helps when
browsing or merging:

```{r childid}
dat <- dat %>%
  group_by(country_iso3, year, HH1, HH2, LN) %>%
  mutate(childid = cur_group_id()) %>%
  ungroup() %>%
  select(childid, everything())

n_distinct(dat$childid) == nrow(dat)
sum(is.na(dat$childid))
```

## Survey design

See [Survey design, weights and standard errors](#c05-survey-design) for the official MICS recipe (`fsweight`, `PSU`, `stratum`). The lab below uses `HH1` as the cluster.

MICS draws children in clusters. Ignoring that design understates uncertainty
and can distort country contrasts. Point estimates use the child weight
`fsweight`; standard errors treat `HH1` as the cluster. For pooled
cross-country work, nest clusters inside each survey.

```{r design-r}
design_one <- function(df) {
  df %>%
    as_survey_design(
          ids = HH1,
      weights = fsweight,
         nest = TRUE
    )
}

design_pooled <- function(df) {
  df %>%
    as_survey_design(
          ids = HH1,
       strata = survey,
      weights = fsweight,
         nest = TRUE
    )
}

# Working sample for most examples: currently enrolled, positive weight
an <- dat %>%
  filter(
    enrolled == 1L,
    !is.na(fsweight), fsweight > 0,
    !is.na(HH1)
  )
```

Rebuild the design after any `mutate()` that creates variables you will
analyse. The design object stores a snapshot of the data.

## Mean age

How old are children currently enrolled in school? Start with Ghana, then scan
the full African set.

```{r mean-age-gha}
gha <- an %>% 
        filter(country_iso3 == "GHA")

design_one(gha) %>%
  summarise(
    age_mean = survey_mean(age, na.rm = TRUE, vartype = "se"),
    n = unweighted(n())
  )
```

Among enrolled children in Ghana 2017, mean age is about **10.4 years**
(SE ≈ 0.09). The FS module covers ages 5–17, so this mean mixes primary and
secondary learners.

```{r mean-age-by-country}
age_by_cty <- design_pooled(an) %>%
  group_by(country_iso3, year) %>%
  summarise(
    age_mean = survey_mean(age, na.rm = TRUE, vartype = NULL),
    n = unweighted(n()),
    .groups = "drop"
  ) %>%
  arrange(country_iso3, year)

age_by_cty
```

Country means sit in a fairly tight band, but age alone is not grade. The next
section fixes the grade and asks how many children are older than they should
be for that grade.

## Over-age children in primary grade 2

Late entry and repetition leave many African classrooms with children far above
the expected age for their grade. If primary starts around age 6, primary grade
2 learners are typically 7 or 8. Ages 10 and above are clearly over-age for
that grade.

Restrict to **primary** (`current_level_h == 1`) and **grade 2**
(`current_grade == 2`). Without the level filter, “grade 2” would also pull in
secondary year 2.

```{r overage-setup}
# Typical ages 7–8; treat age 10+ as clearly over-age for primary grade 2
g2_prim <- an %>%
  filter(
    current_level_h == 1L,
    current_grade == 2L,
    !is.na(age)
  ) %>%
  mutate(overage = as.integer(age >= 10L))

design_one(g2_prim %>% filter(country_iso3 == "GHA")) %>%
  summarise(
    age_mean = survey_mean(age, vartype = "se"),
    overage_share = survey_mean(overage, vartype = "se"),
    n = unweighted(n())
  )
```

In Ghana, mean age in primary grade 2 is about **8.3**, and roughly **19%** of
those children are aged 10 or older.

```{r plot-overage-age, fig.cap="Age distribution in primary grade 2 — Ghana. Ages 10 and above are highlighted as over-age."}
age_gha_g2 <- design_one(g2_prim %>% filter(country_iso3 == "GHA")) %>%
  group_by(age) %>%
  summarise(
    p = survey_mean(vartype = NULL),
    n = unweighted(n()),
    .groups = "drop"
  ) %>%
  mutate(fill = if_else(age >= 10L, aflearn_amber, aflearn_neutral[["context"]]))

age_gha_g2 %>%
  ggplot(aes(x = factor(age), y = p, fill = fill)) +
  geom_col(width = 0.72, colour = NA) +
  scale_fill_identity() +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0.08))
  ) +
  labs(
    title = "Many primary grade 2 learners are older than 7–8",
    subtitle = "Ghana · weighted share by age · amber = age 10+",
    x = "Age",
    y = NULL,
    caption = "Source: MICS6, Ghana 2017."
  ) +
  theme_aflearn() +
  theme(axis.text.x = aflearn_axis_category())
```

The mass sits at 7–8, but the amber bars show a long right tail into the teens.
That is the over-age problem in one classroom label.

```{r overage-by-country}
overage_cty <- design_pooled(g2_prim) %>%
  group_by(country_iso3, year) %>%
  summarise(
    overage_share = survey_mean(overage, vartype = NULL),
    age_mean = survey_mean(age, vartype = NULL),
    n = unweighted(n()),
    .groups = "drop"
  ) %>%
  arrange(desc(overage_share))

overage_cty
```

```{r plot-overage-country, fig.cap="Share of primary grade 2 children aged 10 or older, by survey."}
overage_cty %>%
  mutate(label = paste(country_iso3, year)) %>%
  ggplot(aes(x = overage_share, y = reorder(label, overage_share))) +
  geom_col(width = 0.72, fill = aflearn_electric_blue, colour = NA) +
  scale_x_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0.08))
  ) +
  labs(
    title = "Over-age enrolment in primary grade 2 varies by country",
    subtitle = "Weighted share aged 10+ · enrolled primary grade 2",
    x = "Share of over-age children",
    y = NULL,
    caption = "Source: MIC6. Africa "
  ) +
  theme_aflearn() +
  theme(axis.text.y = aflearn_axis_category())
```

The ranking is stark. Guinea-Bissau and Chad sit near the top because more
than two in five primary grade 2 children are aged 10+. Tunisia and Eswatini sit
near the bottom (about 1–2%). Same grade label, very different age profiles
across this African set.

## Numeracy and reading skills

What do foundational skills look like in one country before we compare school
types? Ghana is a useful case: a large enrolled sample and clear private-school
coding.

```{r mean-outcomes}
# Ghana: mean numeracy among enrolled children with a score
design_one(gha %>% filter(!is.na(numeracy_score))) %>%
  summarise(
    numeracy_mean = survey_mean(numeracy_score, vartype = "se"),
    n = unweighted(n())
  )

# Share meeting foundational reading skills (where assessed)
design_one(gha %>% filter(!is.na(reading_skills))) %>%
  summarise(
    reading_share = survey_mean(reading_skills, vartype = "se"),
    n = unweighted(n())
  )
```

Enrolled Ghanaian children with a numeracy score average about **15 of 21**
items correct (SE ≈ 0.24). Among those with a reading-skills indicator, about
**34%** meet the foundational reading threshold (SE ≈ 0.02). Note that the two samples
differ: numeracy and reading are not always completed for the same children.

```{r plot-numeracy-dist, fig.cap="Distribution of numeracy scores among enrolled Ghanaian children."}


num_share <- design_one(gha %>% filter(!is.na(numeracy_score))) %>%
  group_by(numeracy_score) %>%
  summarise(
    p = survey_mean(vartype = NULL),
    n = unweighted(n()),
    .groups = "drop"
  ) |>
  ungroup()

num_share %>%
   mutate(fill = ifelse(numeracy_score == 21, aflearn_cyan, aflearn_neutral[["context"]])) |> 
  ggplot(aes(x = numeracy_score, y = p, fill = fill)) +
  geom_col(width = 0.9, colour = NA) +
  scale_fill_identity() +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0.08))
  ) +
  labs(
    title = "Numeracy score distribution of Ghanaian children",
    subtitle = "Weighted share · enrolled children with a numeracy score",
    x = "Numeracy score (0–21)",
    y = NULL,
    caption = "Source: MICS6, Ghana 2017."
  ) +
  theme_aflearn()
```

The distribution piles toward the top of the 0–21 scale, with a thinner left
tail of low scorers. That shape is the backdrop for the private-school
contrast.

## Private versus non-private schools

Do privately schooled children score higher on numeracy? Compare weighted
means before adding controls.

```{r means-by-private}
gha_num <- gha %>%
  filter(!is.na(numeracy_score), !is.na(private))

design_one(gha_num) %>%
  group_by(private) %>%
  summarise(
    numeracy_mean = survey_mean(numeracy_score, vartype = "se"),
    n = unweighted(n()),
    .groups = "drop"
  )
```

In Ghana the raw gap is large: about **15.4** among non-private learners versus
**18.5** among private-school learners. That is not a causal estimate. Families
that choose private schools also differ in age mix, gender, and mother’s
education. Regression holds those factors fixed.

## Multivariate regression (Ghana)

Stay in Ghana for the rest of the chapter. A single-country design keeps the
story readable; the same steps extend to other surveys in the file. Restrict to
enrolled children aged 7–14 with non-missing numeracy, private-school status,
sex, and mother’s education (codes 0–5).

```{r ghana-sample}
gha_reg <- gha %>%
  filter(
    age %in% 7:14,
    !is.na(numeracy_score),
    !is.na(private),
    !is.na(female),
    mother_edu_h %in% 0:5
  ) %>%
  mutate(mother_edu_f = factor(mother_edu_h))

des_gha <- design_one(gha_reg)

m1 <- svyglm(numeracy_score ~ private, design = des_gha)
summary(m1)

m2 <- svyglm(
  numeracy_score ~ private + age + female + mother_edu_f,
  design = des_gha
)
summary(m2)
```

In `m1`, the intercept is mean numeracy in non-private schools; the
coefficient on `private` is the associated gap. In `m2`, that gap is
conditional on age, sex, and mother’s education. If the private coefficient
shrinks, part of the raw gap was composition — not an independent school-type
effect.

## Private × female interaction

Does the private-school association differ for girls and boys?

```{r interaction}
m_int <- svyglm(
  numeracy_score ~ private * female + age + mother_edu_f,
  design = des_gha
)
summary(m_int)

coef(m_int)
```

Read the coefficients as follows:

- `private` — association for boys (`female = 0`)
- `female` — girl–boy difference in non-private schools
- `private:female` — how much the private-school association differs for girls

The private-school association for girls is `private + private:female`. A
negative interaction means the private premium is smaller for girls than for
boys (or the reverse if positive). These are associations in Ghana’s MICS
sample, not a multi-country causal claim.

## Predicted means

Coefficients on interactions are easy to misread. Predicted means at the four
private × female cells turn the same model into a table you can plot. Hold age
and mother’s education at the sample profile (modal mother’s education).

```{r predicted-means}
mf <- model.frame(m_int)

mom_mode <- names(sort(table(mf$mother_edu_f), decreasing = TRUE))[1]

newdata <- expand.grid(
  private = c(0, 1),
  female = c(0, 1),
  age = mean(mf$age, na.rm = TRUE),
  mother_edu_f = factor(mom_mode, levels = levels(mf$mother_edu_f))
)

newdata$predicted <- as.numeric(
  predict(m_int, newdata = newdata, type = "response")
)

newdata %>%
  mutate(
    school = if_else(private == 1, "Private", "Non-private"),
    sex_lab = if_else(female == 1, "Girl", "Boy")
  ) %>%
  select(school, sex_lab, age, mother_edu_f, predicted)
```

```{r plot-predicted, fig.cap="Predicted mean numeracy by school type and sex — Ghana, controls at sample profile."}
pred_plot <- newdata %>%
  mutate(
    school = factor(
      if_else(private == 1, "Private", "Non-private"),
      levels = c("Non-private", "Private")
    ),
    sex_lab = factor(
      if_else(female == 1, "Girl", "Boy"),
      levels = c("Boy", "Girl")
    )
  )

cols_sex <- aflearn_pal_categorical(2)
names(cols_sex) <- c("Boy", "Girl")

pred_plot %>%
  ggplot(aes(x = school, y = predicted, fill = sex_lab)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.72, colour = NA) +
  scale_fill_manual(values = cols_sex, name = NULL) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.08))) +
  labs(
    title = "Predicted numeracy by school type and sex",
    subtitle = "Ghana · age and mother’s education held at the sample profile",
    x = NULL,
    y = "Predicted numeracy score",
    caption = "Source: MIC6, Ghana 2017"
  ) +
  theme_aflearn() +
  theme(
    legend.position = "top",
    axis.text.x = aflearn_axis_category()
  )
```

Compare the four bars: the height gap between private and non-private within
each sex is the private association; the gap between girl and boy within each
school type is the gender difference. That is the interaction in one picture.


