## The AFLEARN analysis file {#c03-file}

This is the file `run-it.R` writes: one row per selected child in `fs`. [Files, keys, and who is in fs](#c02-files) sets out who that is. How foundational reading and numeracy are scored is in [Reading and numeracy](#c02-fls). If you do not yet have the file, start from [Getting UNICEF MICS files](#c01-unicef).

``` text
prepare-mics-fs.R
        → Data/UNICEF/<survey>/
harmonize-mics-fs-v2.0.R     spec + crosswalk; already pulls selected hl fields
derive-mics-fl-vars-v1.0.R   passage length, scores, reading_status, max_reading_*
        → mics6-fs-harmonised-v2.0.dta
```

If you only unpacked the zip, `source("Script/run-it.R")` is enough. Re-run the two later scripts when you change the spec:

```{r eval=FALSE}
source("Script/harmonize-mics-fs-v2.0.R")
source("Script/derive-mics-fl-vars-v1.0.R")
```

### Identifiers

| AFLEARN | Source / role |
|---------|----------------|
| `country_iso3`, `year` | Survey |
| `HH1` / `cluster` | Cluster (`HH1`) |
| `HH2` / `hhno` | Household within the cluster |
| `LN` / `linech` | Selected child's line number |
| `sample` | IPUMS `SAMPLE` from the spec |

Do not invent `sample` from country names. Use the value on the file and check it against your extract.

### What is in the file

| Area | Examples |
|------|----------|
| Survey design | `fsweight`, `fshweight` |
| Geography | urban/rural, region |
| Parents | mother and father education |
| Schooling | attendance, education level and grade |
| Child labour | work, chores, hours |
| Parental involvement, functioning, discipline | as fielded |
| Foundational learning | numeracy items; reading scores and `reading_status` |

Source names for every column are in `mics6-crosswalk.xlsx` (also in the [guide repository](https://github.com/datafirst-aflearn/mics-guide/blob/main/Script/mics6-crosswalk.xlsx)). Look up the survey column before you pool a variable.

A shared name does not always mean a shared code. `region`, raw grades, and `school_type` stay country-specific. For education, use the harmonised levels (`current_level_h`, `highest_level_h`, `mother_edu_h`, `father_edu_h`) when you pool. `current_grade` is the year *within* that level: primary year 1 and secondary year 1 both carry the code `1`.

The spec already attaches selected `hl` fields such as parent education and school type. Merge by hand only for variables that are not on this file ([Adding variables from UNICEF files](#c03-merge)).

### What `derive` added

`harmonize-mics-fs-v2.0.R` aligns names. `derive-mics-fl-vars-v1.0.R` then builds the reading fields researchers actually analyse:

| Variable | Meaning |
|----------|---------|
| `passage_language`, `passage_length` | Language and word count of the main passage |
| `words_att`, `words_incorrect` | Words attempted and missed |
| `reading_score`, `reading_accuracy` | Words correct; share of passage length |
| `read_comp_1`–`read_comp_5`, `read_comp_score` | Comprehension items and total |
| `reading_skills` | Official foundational-reading classification |
| `reading_status` | Where the child stopped, or the outcome reached |
| `*B*` / `*C*` and `max_reading_*` | Later passages, and the best result across them |

The official 90% + five-comprehension rule is in [Reading and numeracy](#c02-fls). A missing score is not one kind of missing: children leave the pathway at different points.

### `reading_status`

| Code | Meaning |
|---:|---------|
| 0 | Not interviewed (`interview_result` > 1) |
| 1 | Age outside 7–14 |
| 2 | Caregiver FLS consent not given |
| 3 | Child consent not given |
| 4 | Language mismatch |
| 5 | Child refuses the story |
| 6 | Failed practice |
| 7 | Attempted a passage (`max_reading_score` present) |
| 8 | Met 90% accuracy |
| 9 | Met foundational reading skills |
| 10 | Otherwise |

::: warningbox-header
**Do not recode all non-attempts as children who cannot read**
:::

::: warningbox
A child with no suitable booklet has not demonstrated the same thing as a child who attempted the passage and read no words correctly.
:::

### Checks

::::: {.code-tabs}
:::: {.code-panel data-lang="Stata"}
```stata
use "Data/AFLEARN Harmonised Data/mics6-fs-harmonised-v2.0.dta", clear

tab country_iso3 year
isid country_iso3 year HH1 HH2 LN

tab reading_status if age >= 7 & age <= 14
tab passage_language if country_iso3 == "MWI"
```
::::
:::: {.code-panel data-lang="R"}
```{r eval=FALSE}
library(haven)
library(dplyr)

d <- read_dta("Data/AFLEARN Harmonised Data/mics6-fs-harmonised-v2.0.dta")

d %>% count(country_iso3, year)

d %>%
  count(country_iso3, year, HH1, HH2, LN) %>%
  filter(n > 1)

d %>%
  filter(age >= 7, age <= 14) %>%
  count(reading_status)

d %>%
  filter(country_iso3 == "MWI") %>%
  count(passage_language)
```
::::
:::::

Malawi should show both assessment languages. If the key is not unique, stop before any merge.
