## Attaching the AFLEARN file to IPUMS {#c03-ipums}

Keep the IPUMS child extract for demographic, household, and schooling variables. Attach the AFLEARN file so reading includes every passage. The limit on IPUMS reading variables is in [Getting IPUMS MICS files](#ipums-reading-warning).

The v2 file already carries `sample` (IPUMS `SAMPLE` from the spec).

``` text
AFLEARN                 IPUMS
sample              <-> SAMPLE
cluster             <-> CLUSTER
hhno                <-> HHNO
linech              <-> LINECH
```

::: warningbox-header
**Do not guess IPUMS sample codes from country names**
:::

::: warningbox
Use `sample` on the AFLEARN file and check it against `SAMPLE` in your extract. IPUMS identifiers distinguish survey year and, where needed, national and subnational samples.
:::

Build the extract as children aged 5–17 and keep `SAMPLE`, `CLUSTER`, `HHNO`, and `LINECH`. See [IPUMS MICS linking](https://mics.ipums.org/mics/linking.shtml). Select the AFLEARN columns you need so names do not collide with IPUMS.

### Check uniqueness

::::: {.code-tabs}
:::: {.code-panel data-lang="Stata"}
```stata
use "Data/ipums_mics_fs.dta", clear
duplicates report SAMPLE CLUSTER HHNO LINECH

use "Data/AFLEARN Harmonised Data/mics6-fs-harmonised-v2.0.dta", clear
duplicates report sample cluster hhno linech
```
::::
:::: {.code-panel data-lang="R"}
```{r eval=FALSE}
library(haven)
library(dplyr)

ipums <- read_dta("Data/ipums_mics_fs.dta")
aflearn <- read_dta(
  "Data/AFLEARN Harmonised Data/mics6-fs-harmonised-v2.0.dta"
)

ipums %>%
  count(SAMPLE, CLUSTER, HHNO, LINECH) %>%
  filter(n > 1)

aflearn %>%
  count(sample, cluster, hhno, linech) %>%
  filter(n > 1)
```
::::
:::::

If either check returns duplicates, stop before merging.

### Join

Keep IPUMS as the analysis population. Not every `fs` child has a completed reading assessment; ages 5–6 and 15–17 are outside the FLS target, and eligible children can leave the pathway earlier. The question is whether the child record links, not whether every child has a score.

::::: {.code-tabs}
:::: {.code-panel data-lang="Stata"}
```stata
use "Data/AFLEARN Harmonised Data/mics6-fs-harmonised-v2.0.dta", clear

rename sample SAMPLE
rename cluster CLUSTER
rename hhno HHNO
rename linech LINECH

isid SAMPLE CLUSTER HHNO LINECH

tempfile aflearn
save `aflearn'

use "Data/ipums_mics_fs.dta", clear

merge 1:1 SAMPLE CLUSTER HHNO LINECH using `aflearn'
tab _merge
```
::::
:::: {.code-panel data-lang="R"}
```{r eval=FALSE}
library(haven)
library(dplyr)

ipums <- read_dta("Data/ipums_mics_fs.dta")
aflearn <- read_dta(
  "Data/AFLEARN Harmonised Data/mics6-fs-harmonised-v2.0.dta"
)

analysis <- ipums %>%
  left_join(
    aflearn,
    by = c(
      "SAMPLE" = "sample",
      "CLUSTER" = "cluster",
      "HHNO" = "hhno",
      "LINECH" = "linech"
    ),
    relationship = "one-to-one"
  )
```
::::
:::::

Do not drop unmatched records until you know why they did not match. Once AFLEARN reading is on the file, drop or rename the IPUMS reading variables so you do not use them by accident.

### After the merge

Use the IPUMS age variable from your extract (often `AGE`). Malawi must show both assessment languages.

::::: {.code-tabs}
:::: {.code-panel data-lang="Stata"}
```stata
tab SAMPLE
tab reading_status if AGE >= 7 & AGE <= 14
tab SAMPLE passage_language
```
::::
:::: {.code-panel data-lang="R"}
```{r eval=FALSE}
analysis %>%
  count(SAMPLE)

analysis %>%
  filter(AGE >= 7, AGE <= 14) %>%
  count(SAMPLE, reading_status)

analysis %>%
  count(SAMPLE, passage_language)
```
::::
:::::

If the study uses only single-passage countries, the AFLEARN reading fields are still the simpler common definition: the same `reading_status`, passage length, and accuracy rule in every supported survey.
