
# Harmonising MICS6 reading outcomes {#c05-harmonize}

Cross-country work on foundational reading needs comparable outcomes: word
scores, comprehension, and skills indicators pooled across MICS6 surveys.
Official country files use different names and layouts, so they do not pool
cleanly on their own.

[IPUMS MICS](https://mics.ipums.org/mics/) solves much of that problem. It
harmonises variable names and codes and releases integrated extracts that
serve most cross-country analyses well.

It has one decisive limit for reading. Where a survey administered several
passages in different languages, the IPUMS extract keeps only the first
passage. In multilingual countries a child's strongest or language-matched
attempt may sit on a later passage, so those extracts are a poor base for
reading-achievement analysis.

This chapter builds the reading outcomes from the official per-survey
`fs.sav` files prepared in [Download the MICS Data](#c01-download). It
harmonises names, retains every passage (and folds B/C where the scripts
require), derives passage length from the data, builds scores and skills on
a common footing, and writes `data/mics6_reading_harmonized.dta` with merge
keys so you can still link background variables from an IPUMS extract.

One complete **v1.1** R script accompanies the chapter. It lives in `data/`
next to `MICS_Datasets/`:

| Script | Language |
|---|---|
| `data/harmonize-mics-reading-v1.1.R` | R, using the tidyverse |

Download the file from AFLEARN's GitHub page, set the working directory to the
`data/` folder, and run:

```{r eval=FALSE}
source("data/harmonize-mics-reading-v1.1.R")
```

It writes `data/mics6_reading_harmonized.dta`. The sections below explain what
each stage does and why, so you can adapt the logic rather than treat it as a
black box.

## Before you start

You need `data/MICS_Datasets/<SURVEY>/fs.sav` in place from
[Download the MICS Data](#c01-download) (`prepare-mics-fs.R`). The
`fs.sav` file holds the Foundational Learning Skills questionnaire,
administered to children aged 5–17, with the reading assessment itself
targeted at ages 7–14. The guide workspace looks like this:

``` text
mics-guide/
└── data/
    ├── MICS_Datasets/
    │   └── BEN_2021_MICS6_v01_M/fs.sav
    └── harmonize-mics-reading-v1.1.R
```

Currently, 18 country codes are supported:

``` text
BEN CAF COD COM GHA GMB GNB LSO MDG MWI NGA SLE STP SWZ TCD TGO TUN ZWE
```

Any other survey folder in `data/MICS_Datasets/` is skipped with a message.

## The harmonisation pipeline at a glance

``` text
for each survey folder
  1.  identify the survey from its folder name
  2.  fix questionnaire numbering quirks
  3.  harmonise variable names 
  4.  repair known data problems (SLE swap; early refusal cleaning)
  5.  assign passage language
  6.  derive passage length from the data
  7.  Move Malawi's Chichewa-only attempts into the main slots
  8.  build reading, comprehension, and practice scores
  9.  build accuracy and foundational skills
  10. classify the assessment outcome (codes 0–9)
  11. add merge keys and keep the analysis variables
append all surveys, then label and save
```

## Identify the survey

Folder names carry an ISO3 country code and a year, as in
`BEN_2021_MICS6_v01_M`. A regular expression pulls both out. Anything that
does not match, is not in the supported list, or has no `fs.sav`, is skipped.

```{r eval=FALSE}
for (folder in list.dirs(root, recursive = FALSE, full.names = FALSE)) {
  m <- regmatches(folder, regexec("^([A-Za-z]{3})_([0-9]{4})_", folder))[[1]]
  if (length(m) == 0) next
  iso  <- toupper(m[2])
  year <- m[3]
  if (!iso %in% supported) next

  d <- read_sav(file.path(root, folder, "fs.sav")) %>% zap_labels()
}
```

`zap_labels()` strips the SPSS value labels but keeps the
underlying numeric codes. Fresh labels are applied at the very end.

## Fix questionnaire numbering

Two surveys need repairs before renaming. Zimbabwe stores five comprehension
items in `FL22B` through `FL22F`. Comoros shifts `FL21BE`/`FL21BF` by one
position (it does **not** rename the whole `FL21B*` series onto `FL22*`).

```{r eval=FALSE}
if (iso == "ZWE") {
  d <- d %>% rename(FL22D = "FL22E", FL22E = "FL22F")
}
if (iso == "COM") {
  d <- d %>% rename(FL21BD = "FL21BE", FL21BE = "FL21BF")
}
```

These renames are deliberately not wrapped in `capture`. If the data ever
changes shape, you want a loud error here.

## Harmonise variable names

A slim set of renames maps questionnaire codes onto readable names used
downstream. Each rename is conditional: a survey that never asked a question
keeps its own variables and the script carries on. A `B` or `C` suffix marks
the second or third reading passage.

```{r eval=FALSE}
d <- cap_rename(d, c(
  age = "CB3", consent = "FL1", child_consent = "FL3",
  likestory = "FL10", words_att = "FL20A", words_incorrect = "FL20B"
  # ... full map in data/harmonize-mics-reading-v1.1.R
))
```

### Chad school language

For Chad only, `FL9A` can supply `lang_school`, and non-missing `FL9B` values
are copied in afterwards:

```{r eval=FALSE}
if (iso == "TCD") {
  d <- cap_rename(d, c(lang_school = "FL9A"))
  d <- d %>% mutate(
    lang_school = if_else(present(FL9B), FL9B, lang_school)
  )
}
```

### Country-specific comprehension renames

Most surveys store passage-1 comprehension in `FL22A`–`FL22E`. Four surveys
store passage 1 in `FL21BA`–`FL21BE` and passage 2 in `FL22A`–`FL22E`:

| Surveys | Passage 1 | Passage 2 |
|---|---|---|
| Default (everyone else) | `FL22A`–`FL22E` → `read_comp_1`–`5` | other codes (`FL122*`, `FLB22*`, …) |
| SWZ, NGA, ZWE, COM | `FL21BA`–`FL21BE` → `read_comp_1`–`5` | `FL22A`–`FL22E` → `read_compB_1`–`5` |

COM is excluded from the default `FL22` rename so those items remain available
for passage B.

After renaming, the script guarantees that the variables the pipeline needs
exist, creating them as missing where a survey omitted them (`ensure`).

### A note on missing values

In R, comparisons with `NA` yield `NA`, not `FALSE`. Helpers make the
intended tests explicit:

```{r eval=FALSE}
present <- function(x) !is.na(x)
eq      <- function(x, v) !is.na(x) & x == v
ne_obs  <- function(x, v) !is.na(x) & x != v
```

## Repair known data problems

### Sierra Leone word swap

Some Sierra Leone records have words-attempted and words-incorrect stored the
wrong way round. The fix swaps them where attempted is smaller than incorrect,
and only where the incorrect count is actually recorded (not missing).
A careless “less than or equal to missing” style check would create false zeros.

### Early refusal cleaning

In São Tomé and Príncipe, Madagascar, and the Central African Republic
(`CAF`, not `CAR`), children who refused the story or failed practice can
still have non-missing word counts. Those counts are cleared before
scores are built:

```{r eval=FALSE}
if (iso %in% c("STP", "MDG", "CAF")) {
  drop1 <- !eq(d$likestory, 1) | ne_obs(d$practice_correct, 1)
  d <- d %>% mutate(
    words_att       = if_else(drop1, NA_real_, words_att),
    words_incorrect = if_else(drop1, NA_real_, words_incorrect)
  )
}
```

Madagascar applies the same idea to passages B and C using `FL110` / `FL210`
and the practice-B/C items.

## Passage language

Language is assigned next, before length. Languages for the single-language surveys are English (Ghana, Sierra Leone, Gambia), French (Benin, Central African Republic, Chad, Comoros, DR Congo, Togo), Arabic (Tunisia) and Portuguese (São Tomé, Guinea-Bissau). For these countries, the language is assigned to the passage language variable. 

| Group | Code | Surveys |
|---|---|---|
| English | 1010 | GHA, SLE, GMB |
| French | 1020 | BEN, CAF, TCD, COM, COD, TGO |
| Portuguese | 1040 | STP, GNB |
| Arabic | 2010 | TUN |

For Nigeria, Eswatini, Lesotho, Madagascar, Malawi, and Zimbabwe, the variables used to identify the passage language differ across surveys. The harmonisation process takes this into account (see the scripts). For instance, Zimbabwe is the most involved, because the passage language has to be inferred from the language of instruction, then from the language spoken at home, then from the child's own choice of story 


```{r eval = FALSE}
if (iso == "ZWE") {
  d <- d %>% ensure(c("FL10C", "FL21D")) %>% mutate(
    passage_language = case_when(
      eq(lang_school, 1) ~ 1010,
      eq(lang_school, 2) ~ 3140,
      eq(lang_school, 3) ~ 3150,
      present(lang_school) & lang_school >= 7 & lang_school <= 9 ~ NA_real_,
      TRUE ~ as.numeric(lang_school)
    ),
    passage_language = if_else(is.na(words_att), NA_real_, passage_language),
    passage_language = case_when(
      eq(lang_home, 1) & is.na(passage_language) & present(words_att) ~ 1010,
      eq(lang_home, 2) & is.na(passage_language) & present(words_att) ~ 3140,
      eq(lang_home, 3) & is.na(passage_language) & present(words_att) ~ 3150,
      TRUE ~ passage_language
    ),
    passage_language = case_when(
      eq(FL10C, 1) & present(words_att) ~ 1010,
      eq(FL10C, 2) & present(words_att) ~ 3140,
      eq(FL10C, 3) & present(words_att) ~ 3150,
      TRUE ~ passage_language
    ),
    passage_length  = if_else(present(words_att), 72, NA_real_),
    passageB_length = if_else(present(wordsB_att), 62, NA_real_)
  )
}
```


## Passage length from the data

For each language, passage length is the maximum
words attempted among children who read that language:

\[
\text{passage_length} =
\max(\text{words_att} \mid \text{passage_language})
\]

```{r eval=FALSE}
d$passage_length <- passage_len_by_lang(d$words_att, d$passage_language)
# same idea for B/C when those passages exist
```

This puts different languages on a common accuracy scale without maintaining a
country-by-country word-count table. It assumes that, within a language, at
least one child reached (or nearly reached) the end of the passage. If every
child stopped early, the estimated length would be too short and accuracy
could exceed 1.

## Chichewa-only attempts

In Malawi, English is passage A and Chichewa is passage B. Some children
attempt only Chichewa. Those records are moved into the main slots so the
pooled file treats them as the child’s primary attempt, and the B slots are
cleared:

```{r eval=FALSE}
if (iso == "MWI") {
  tmp <- is.na(d$words_att) & present(d$wordsB_att)
  d <- d %>% mutate(
    words_att        = if_else(tmp, wordsB_att, words_att),
    words_incorrect  = if_else(tmp, wordsB_incorrect, words_incorrect),
    passage_language = if_else(tmp, passageB_language, passage_language),
    passage_length   = if_else(tmp, passageB_length, passage_length)
    # ... comprehension items, then blank B
  )
}
```

## Reading, comprehension, and practice scores

With cleaned word counts in place:

\[
\text{reading_score} = \text{words_att} - \text{words_incorrect}
\]

Comprehension scores count how many of the five items are coded 1 (correct),
and are missing when the first item was never asked. Practice is passed only
if the practice sentence and both practice questions are correct.
`fail_practice` is set if any passage’s practice was failed.

```{r eval=FALSE}
d <- d %>% mutate(reading_score = words_att - words_incorrect)
d$read_comp_score <- comp_score(d, "read_comp")
d$practice_outcome <- practice_outcome(
  d, "practice_correct", "practice_question1", "practice_question2"
)
d$fail_practice <- 1 - d$practice_outcome
```

## Accuracy and foundational reading skills

### Accuracy

**Accuracy** is the share of the passage the child read correctly:

\[
\text{reading_accuracy} = \frac{\text{reading_score}}{\text{passage_length}}
\]

Dividing by `passage_length` puts different languages on a common scale. The
same formula is used for a second passage as `readingB_accuracy` when that
passage exists.

### Foundational reading skills

**Foundational reading skills** is 1 only when *both* hold (among children
with a reading score):

1.  All five comprehension questions are correct:
    \(\text{read_comp_score} = 5\).
2.  At least 90% of the passage words are correct:

\[
\text{cutoff} = \lfloor 0.9 \times \text{passage_length} \rfloor
\]

\[
\text{reading_skills} =
\begin{cases}
1 & \text{if } \text{read_comp_score} = 5
  \text{ and } \text{reading_score} \ge \text{cutoff} \\
0 & \text{otherwise (when } \text{reading_score} \text{ is non-missing)}
\end{cases}
\]

```{r eval=FALSE}
d <- d %>% mutate(
  reading_accuracy = reading_score / passage_length,
  cutoff = trunc(0.9 * passage_length),
  reading_skills = if_else(
    present(reading_score),
    as.numeric(eq(read_comp_score, 5) & ge_obs(reading_score, cutoff)),
    NA_real_
  )
)
```


## Classify the assessment outcome

`reading_status` records why a child does or does not have a reading score.
Codes 2–4 only fill gaps; codes 5–8 overwrite; code **9** catches anyone still
unclassified (inconsistent data, for example comprehension without word
counts):

| Code | Meaning |
|---|---|
| 0 | Not interviewed |
| 1 | Age under 7 or over 14 |
| 2 | Caregiver consent not given |
| 3 | Child consent not given |
| 4 | Assessment language does not match the child's language |
| 5 | Child does not want to read the story |
| 6 | Failed the practice sentence and questions |
| 7 | Attempted a reading passage |
| 8 | Read 90% of words and answered all comprehension questions correctly |
| 9 | Not classified — data inconsistent |

```{r eval=FALSE}
d <- d %>% mutate(
  reading_status = if_else(present(interview_result) & interview_result > 1, 0, NA_real_),
  reading_status = if_else(present(age) & (age < 7 | age > 14), 1, reading_status),
  reading_status = if_else(ne_obs(consent, 1) & is.na(reading_status), 2, reading_status),
  reading_status = if_else(ne_obs(child_consent, 1) & is.na(reading_status), 3, reading_status),
  reading_status = if_else(eq(lang_mismatch, 1) & is.na(reading_status), 4, reading_status),
  reading_status = if_else(eq(child_refuses_read, 1), 5, reading_status),
  reading_status = if_else(eq(fail_practice, 1), 6, reading_status),
  reading_status = if_else(present(max_reading_score), 7, reading_status),
  reading_status = if_else(eq(max_reading_skills, 1), 8, reading_status),
  reading_status = if_else(is.na(reading_status), 9, reading_status)
)
```

## Merge keys and variable selection

Each survey is uniquely identified by cluster, household and line number.
Those are copied to the IPUMS-MICS names `cluster`, `hhno`, and `linech`.
Country and year are added, then the file is cut to a **slim** analysis set:

```{r eval=FALSE}
d %>% select(
  any_of(c("country_iso3", "year", "cluster", "hhno", "linech",
           "HH1", "HH2", "LN", "reading_status")),
  matches("^reading"), matches("^read_comp"), matches("^practice"),
  matches("^passage"), matches("^words.*_att$"), matches("^words.*_incorrect$")
)
```


## What the harmonized dataset contains

One row per child in the Foundational Learning Skills module, across all
supported surveys.

| Variable | Meaning |
|---|---|
| `country_iso3`, `year` | Survey identification |
| `cluster`, `hhno`, `linech` | Merge keys for IPUMS-MICS (copies of `HH1`, `HH2`, `LN`) |
| `reading_status` | Outcome of the reading assessment, coded 0–9 |
| `words_att`, `words_incorrect` | Raw word counts for passage 1 |
| `reading_score` | Words read correctly, passage 1 |
| `reading_accuracy` | `reading_score / passage_length` |
| `reading_skills` | 90% of words plus all comprehension correct |
| `read_comp_1`–`read_comp_5`, `read_comp_score` | Comprehension items and total |
| `practice_*`, `practice_outcome` | Practice sentence and questions |
| `passage_language`, `passage_length` | Passage characteristics (length from data) |
| `*B_*` equivalents | The same measures for the second passage |
| `max_reading_score`, `max_reading_accuracy`, `max_reading_skills` | Best result across passages |

## Known limitations

- Data-driven passage length uses the maximum words attempted within each
  language. If nobody in that language reached the end of the passage, the
  estimated length is too short and accuracy can exceed 1.
- Year differences within a country are not modelled. Tunisia appears
  twice (2018 and 2023) and both use the same language code.


## Checks after running

```{r eval=FALSE}
library(tidyverse)
library(haven)

d <- read_dta("data/mics6_reading_harmonized.dta")

count(d, country_iso3, year)
count(d, reading_status)
summary(d$reading_accuracy)
d %>% filter(reading_status == 8) %>% nrow()
```

Accuracy should sit between 0 and 1. Values above 1 point to an underestimated
passage length for that language.

## Next steps

With `data/mics6_reading_harmonized.dta` in place:

- Follow [R Tutorial](#c03-r) to analyse it in R.
- Follow [Stata Tutorial](#c04-stata) to analyse it in Stata.
- Build non-reading FS themes (schooling, labour, discipline, numeracy setup,
  and more) with [Harmonising MICS6 FS background themes](#c06-harmonize-fs)
  and merge on `country_iso3`, `year`, and `HH1`/`HH2`/`LN`.

Because the file carries `cluster`, `hhno` and `linech`, it can also be merged
onto an IPUMS-MICS extract to combine these reading outcomes with harmonised
household and background variables.
