# Harmonising MICS6 reading outcomes {#c05-harmonize}

MICS6 asked children the same reading questions in every country, but each
national questionnaire numbered them differently. One survey stores the
comprehension items as `FL22A`–`FL22E`, another as `FL21BA`–`FL21BF`, a third
as `FLB22A`–`FLB22E`. Some countries administered one reading passage, others
two or three. Passages differ in language and in length, so a raw word count
means different things in different countries.

This chapter walks through building a single comparable dataset of
**foundational reading outcomes** from the per-survey `fs.sav` files you
prepared in [Download the MICS Data](#c01-download). By the end you will have
one file covering every supported survey, with harmonised variable names,
comparable reading measures, and a single classification of what happened in
each child's reading assessment.

Two complete scripts accompany the chapter. They implement identical
definitions, so use whichever language you work in:

| Script | Language |
|---|---|
| `do/harmonize-mics-reading.do` | Stata 16 or later |
| `R/harmonize-mics-reading.R` | R, using the tidyverse |

Run either one from the project root. Throughout the chapter, code is shown
in pairs labelled **Stata** then **R**, so you can follow in either language.

<p class="code-lang code-lang-stata">Stata</p>

```{stata eval=FALSE}
do "do/harmonize-mics-reading.do"
```

<p class="code-lang code-lang-r">R</p>

```{r eval=FALSE}
source("R/harmonize-mics-reading.R")
```

Both write `data/mics6_reading_harmonized.dta`. The sections below explain what each stage does and why, so you can adapt the logic rather than treat it as a black box.

## Before you start

You need `data/MICS_Datasets/<SURVEY>/fs.sav` in place, which is what
`R/prepare-mics-fs.R` produces. The `fs.sav` file holds the Foundational
Learning Skills questionnaire, administered to children aged 5–17, with the
reading assessment itself targeted at ages 7–14.

Eighteen country codes are supported, because the passage language and length
rules below were written for them:

``` text
BEN CAF COD COM GHA GMB GNB LSO MDG MWI NGA SLE STP SWZ TCD TGO TUN ZWE
```

## The pipeline at a glance

The scripts process one survey at a time, then append the results:

``` text
for each survey folder
  1.  identify the survey from its folder name
  2.  fix questionnaire numbering quirks
  3.  harmonise variable names
  4.  repair known data problems
  5.  build reading scores
  6.  build comprehension scores
  7.  build practice outcomes
  8.  attach passage language and length
  9.  build accuracy and foundational skills
  10. collapse extra passages where needed
  11. classify the assessment outcome
  12. add merge keys and keep the analysis variables
append all surveys, then label and save
```

## Identify the survey

Folder names carry the information needed to label each survey: an ISO3
country code and a year, as in `BEN_2021_MICS6_v01_M`. A regular expression
pulls both out. Anything that does not match, is not in the supported list, or
has no `fs.sav`, is skipped with a message rather than stopping the run.

<p class="code-lang code-lang-stata">Stata</p>

```{stata eval=FALSE}
local countries : dir "`root'" dirs "*"

foreach c of local countries {
    if !regexm("`c'", "^([A-Za-z]{3})_([0-9]{4})_") {
        display as text "skip `c': cannot read ISO3 code and year from folder name"
        continue
    }
    local iso  = upper(regexs(1))
    local year = regexs(2)

    if strpos(" `supported' ", " `iso' ") == 0 {
        display as text "skip `c': `iso' is not covered by the passage rules"
        continue
    }

    import spss using "`root'/`c'/fs.sav", clear
```

<p class="code-lang code-lang-r">R</p>

```{r eval=FALSE}
for (folder in list.dirs(root, recursive = FALSE, full.names = FALSE)) {
  m <- regmatches(folder, regexec("^([A-Za-z]{3})_([0-9]{4})_", folder))[[1]]
  if (length(m) == 0) next
  iso  <- toupper(m[2])
  year <- m[3]
  if (!iso %in% supported) next

  d <- read_sav(file.path(root, folder, "fs.sav")) %>% zap_labels()
}
```

`zap_labels()` in the R version strips the SPSS value labels but keeps the
underlying numeric codes. That matters: every comparison later in the pipeline
is written against numeric codes, exactly as Stata sees them after
`import spss`. Fresh labels are applied at the very end.

## Fix questionnaire numbering

Two surveys need repairs before any renaming, because their comprehension
items are shifted by one position. Zimbabwe stores five items in `FL22B`
through `FL22F`, and Comoros stores its series under `FL21B*` rather than
`FL22*`.

<p class="code-lang code-lang-stata">Stata</p>

```{stata eval=FALSE}
if "`iso'" == "ZWE" {
    rename FL22E  FL22D
    rename FL22F  FL22E
}
if "`iso'" == "COM" {
    rename FL21BE FL21BD
    rename FL21BF FL21BE
    rename FL21B? FL22?
}
```

<p class="code-lang code-lang-r">R</p>

```{r eval=FALSE}
if (iso == "ZWE") {
  d <- d %>% rename(FL22D = "FL22E", FL22E = "FL22F")
}
if (iso == "COM") {
  d <- d %>% rename(FL21BD = "FL21BE", FL21BE = "FL21BF")
  b_cols <- grep("^FL21B[A-Z]$", names(d), value = TRUE)
  names(d)[match(b_cols, names(d))] <- sub("^FL21B", "FL22", b_cols)
}
```

These renames are deliberately not wrapped in `capture`. If the data ever
changes shape, you want a loud error here rather than a silent misalignment of
comprehension items.

## Harmonise variable names

This is the heart of the exercise: roughly 65 renames that map questionnaire
codes onto readable names. The key is that each rename is *conditional*. A
survey that never asked a question keeps its own variables and the script
carries on. Where several codes mean the same thing across countries, they
point at the same target name and the first match wins.

A `B` or `C` suffix marks the second or third reading passage.

<p class="code-lang code-lang-stata">Stata</p>

```{stata eval=FALSE}
capture rename CB3    age
capture rename FL1    consent
capture rename FL3    child_consent
capture rename FL10   likestory
capture rename FL20A  words_att
capture rename FL20B  words_incorrect
capture rename FL22A  read_comp_1
* ... through read_comp_5, then the B and C passage equivalents
capture rename FL9    lang_school
capture rename FL9A   lang_school     // alternative code, only used if FL9 was absent
```

In R the same idea is a named vector plus a small helper. Duplicate target
names are allowed and are tried in order:

<p class="code-lang code-lang-r">R</p>

```{r eval=FALSE}
rename_map <- c(
  age             = "CB3",
  consent         = "FL1",
  child_consent   = "FL3",
  likestory       = "FL10",
  words_att       = "FL20A",
  words_incorrect = "FL20B",
  read_comp_1     = "FL22A",
  lang_school     = "FL9",
  lang_school     = "FL9A"
  # ... full map in R/harmonize-mics-reading.R
)

apply_rename_map <- function(df, map = rename_map) {
  for (i in seq_along(map)) {
    new <- names(map)[i]
    old <- unname(map[i])
    if (old %in% names(df) && !(new %in% names(df))) {
      names(df)[names(df) == old] <- new
    }
  }
  df
}
```

After renaming, both scripts guarantee that the variables the pipeline needs
exist, creating them as missing where a survey omitted them. This replaces
dozens of existence checks further down and prints a note each time, so gaps
stay visible:

<p class="code-lang code-lang-stata">Stata</p>

```{stata eval=FALSE}
ensure `required'     // age consent child_consent likestory words_att ...
```

<p class="code-lang code-lang-r">R</p>

```{r eval=FALSE}
d <- d %>% ensure(required_vars)
```

### A note on missing values

Stata treats missing as larger than any number, and comparisons involving
missing usually evaluate to false rather than to missing. R does neither. R makes
those rules explicit with three small helpers, used everywhere below:

<p class="code-lang code-lang-r">R</p>

```{r eval=FALSE}
present <- function(x) !is.na(x)              # Stata: x < .
eq      <- function(x, v) !is.na(x) & x == v  # Stata: x == v
ne_obs  <- function(x, v) !is.na(x) & x != v  # Stata: x != v & x < .
```

Reading `present()`, `eq()` and `ne_obs()` as "non-missing", "equals" and
"non-missing and not equal to" makes the rest of the R code map line by line
onto the Stata.

## Repair known data problems

In Sierra Leone some records have words-attempted and words-incorrect stored
the wrong way round. The fix swaps them where attempted is smaller than
incorrect, and only where the incorrect count is actually recorded.

<p class="code-lang code-lang-stata">Stata</p>

```{stata eval=FALSE}
if "`iso'" == "SLE" {
    generate tmp  = words_att       if words_att < words_incorrect & words_incorrect < .
    generate tmp2 = words_incorrect if words_att < words_incorrect & words_incorrect < .
    replace words_att       = tmp2 if tmp2 < .
    replace words_incorrect = tmp  if tmp  < .
    drop tmp tmp2
}
```

<p class="code-lang code-lang-r">R</p>

```{r eval=FALSE}
if (iso == "SLE") {
  swap <- present(d$words_incorrect) & present(d$words_att) &
    d$words_att < d$words_incorrect
  d <- d %>% mutate(
    words_att       = if_else(swap, d$words_incorrect, words_att),
    words_incorrect = if_else(swap, d$words_att, words_incorrect)
  )
}
```



## Reading scores

The reading score is simply the number of words a child read correctly:
words attempted less words incorrect or missed. Second and third passages get
the same treatment where they exist.

<p class="code-lang code-lang-stata">Stata</p>

```{stata eval=FALSE}
generate reading_score = words_att - words_incorrect

capture confirm variable wordsB_att
if _rc == 0 generate readingB_score = wordsB_att - wordsB_incorrect
```

<p class="code-lang code-lang-r">R</p>

```{r eval=FALSE}
d <- d %>% mutate(reading_score = words_att - words_incorrect)

if ("wordsB_att" %in% names(d)) {
  d <- d %>% mutate(readingB_score = wordsB_att - wordsB_incorrect)
}
```

Three surveys need an extra correction. In São Tomé and Príncipe, Madagascar
and the Central African Republic, children who refused to read or who failed
the practice sentence still have non-missing word counts, which would otherwise
be read as genuine (and very low) scores. Those records are set to missing:

<p class="code-lang code-lang-stata">Stata</p>

```{stata eval=FALSE}
if "`iso'" == "STP" | "`iso'" == "CAF" {
    replace reading_score = . if likestory != 1 | (practice_correct != 1 & practice_correct < .)
    replace words_att       = . if reading_score == .
    replace words_incorrect = . if reading_score == .
}
```

<p class="code-lang code-lang-r">R</p>

```{r eval=FALSE}
if (iso %in% c("STP", "CAF", "MDG")) {
  drop1 <- !eq(d$likestory, 1) | ne_obs(d$practice_correct, 1)
  d <- d %>% mutate(
    reading_score   = if_else(drop1, NA_real_, reading_score),
    words_att       = if_else(drop1, NA_real_, words_att),
    words_incorrect = if_else(drop1, NA_real_, words_incorrect)
  )
}
```

Note the asymmetry in the condition, which is intentional: `likestory != 1`
also catches children for whom the question is missing, while the practice
condition applies only where a practice result was recorded.

## Comprehension scores

Each passage has five comprehension questions, coded 1 for a correct answer.
The score counts how many of the five are correct. It is deliberately missing,
rather than zero, when the first item was never asked, so children who never
reached the comprehension questions are not recorded as scoring zero.

<p class="code-lang code-lang-stata">Stata</p>

```{stata eval=FALSE}
generate read_comp_score = 0 if read_comp_1 < .
foreach var of varlist read_comp_? {
    replace read_comp_score = read_comp_score + 1 if `var' == 1
}
```

<p class="code-lang code-lang-r">R</p>

```{r eval=FALSE}
comp_score <- function(df, prefix) {
  first <- paste0(prefix, "_1")
  if (!first %in% names(df)) return(NULL)
  items <- intersect(paste0(prefix, "_", 1:5), names(df))
  correct <- Reduce(`+`, lapply(items, function(v) as.integer(eq(df[[v]], 1))))
  if_else(is.na(df[[first]]), NA_real_, as.numeric(correct))
}

d$read_comp_score <- comp_score(d, "read_comp")
```

## Practice outcomes

Before reading a passage, each child reads a practice sentence and answers two
practice questions. A child passes only if all three are correct.

<p class="code-lang code-lang-stata">Stata</p>

```{stata eval=FALSE}
generate practice_outcome = practice_correct == 1 &     ///
                            practice_question1 == 1 &   ///
                            practice_question2 == 1 if practice_correct < .

generate fail_practice = 1 - practice_outcome
```

<p class="code-lang code-lang-r">R</p>

```{r eval=FALSE}
practice_outcome <- function(df, correct, q1, q2) {
  if_else(is.na(df[[correct]]), NA_real_,
          as.numeric(eq(df[[correct]], 1) & eq(df[[q1]], 1) & eq(df[[q2]], 1)))
}

d$practice_outcome <- practice_outcome(d, "practice_correct",
                                       "practice_question1", "practice_question2")
d$fail_practice <- 1 - d$practice_outcome
```

`fail_practice` is then set to 1 if the practice for *any* later passage was
failed. It feeds the outcome classification in Step 11, where a child who went
on to attempt a passage is reclassified anyway.

## Passage language and length

This is the largest country-specific block, and the reason the supported
country list exists. Reading accuracy is a proportion, so every survey needs a
denominator: the number of words in the passage the child read. Multilingual
surveys read the language from a survey variable and assign the matching word
count; single-language surveys assign one length directly.

| Survey | Rule |
|---|---|
| NGA | Recodes `lang1`/`lang2`; passage 1 lengths 72 (English), 76 (Hausa), 88 (Igbo), 81 (Yoruba) |
| SWZ | Recodes `langS1`/`langS2`; 74 (English), 41 (Siswati) |
| LSO | Recodes `FL100`; 64 (Sesotho), 71 (English); passages 2 and 3 fixed |
| MDG | Recodes `FL100`; 84 (Malagasy), 64 (French); passages 2 and 3 fixed |
| MWI | English 61, Chichewa 74 |
| ZWE | `lang_school`, falling back to `lang_home` then `FL10C`; 72 and 62 |
| TUN, SLE, GNB, GMB | 72 |
| BEN, TCD, COM, TGO | 81 |
| GHA | 69 |
| CAF | 87 |
| COD | 85 |
| STP | 76 |

Languages for the single-language surveys are English (Ghana, Sierra Leone,
Gambia), French (Benin, Central African Republic, Chad, Comoros, DR Congo,
Togo) and Portuguese (São Tomé, Guinea-Bissau).

<p class="code-lang code-lang-stata">Stata</p>

```{stata eval=FALSE}
if inlist("`iso'", "TUN", "SLE", "GNB", "GMB") generate passage_length = 72 if words_att < .
if inlist("`iso'", "BEN", "TCD", "COM", "TGO") generate passage_length = 81 if words_att < .
if "`iso'" == "GHA" generate passage_length = 69 if words_att < .

if inlist("`iso'", "GHA", "SLE", "GMB") generate passage_language = 1010 if reading_score < .
```

<p class="code-lang code-lang-r">R</p>

```{r eval=FALSE}
one_length <- c(TUN = 72, SLE = 72, GNB = 72, GMB = 72,
                BEN = 81, TCD = 81, COM = 81, TGO = 81,
                GHA = 69, CAF = 87, COD = 85, STP = 76)

if (iso %in% names(one_length)) {
  len <- unname(one_length[iso])
  d <- d %>% mutate(passage_length = if_else(present(words_att), len, NA_real_))
}
```

The multilingual blocks follow the same shape. Zimbabwe is the most involved,
because the passage language has to be inferred from the language of
instruction, then from the language spoken at home, then from the child's own
choice of story:

<p class="code-lang code-lang-r">R</p>

```{r eval=FALSE}
if (iso == "ZWE") {
  d <- d %>% ensure(c("FL10C", "FL21D")) %>% mutate(
    passage_language = case_when(
      eq(lang_school, 1) ~ 1010,
      eq(lang_school, 2) ~ 3140,
      eq(lang_school, 3) ~ 3150,
      present(lang_school) & lang_school >= 7 & lang_school <= 9 ~ NA_real_,
      TRUE ~ as.numeric(lang_school)
    ),
    passage_language = if_else(is.na(words_att), NA_real_, passage_language),
    passage_language = case_when(
      eq(lang_home, 1) & is.na(passage_language) & present(words_att) ~ 1010,
      eq(lang_home, 2) & is.na(passage_language) & present(words_att) ~ 3140,
      eq(lang_home, 3) & is.na(passage_language) & present(words_att) ~ 3150,
      TRUE ~ passage_language
    ),
    passage_language = case_when(
      eq(FL10C, 1) & present(words_att) ~ 1010,
      eq(FL10C, 2) & present(words_att) ~ 3140,
      eq(FL10C, 3) & present(words_att) ~ 3150,
      TRUE ~ passage_language
    ),
    passage_length  = if_else(present(words_att), 72, NA_real_),
    passageB_length = if_else(present(wordsB_att), 62, NA_real_)
  )
}
```

The `TRUE ~ as.numeric(lang_school)` fallback is not decoration. Stata's
`recode` leaves unlisted codes unchanged, so the R version copies the original
value through for anything the recode does not name.

## Accuracy and foundational reading skills

With a denominator in place, two measures follow. They answer different
questions: how well the child read the passage, and whether that reading meets
a threshold that also requires full comprehension.

### Accuracy

Accuracy is the share of the passage the child read correctly. It is a
continuous measure between 0 and 1 (when the passage length is correct):

\[
\text{reading_score} = \text{words_att} - \text{words_incorrect}
\]

\[
\text{reading_accuracy} = \frac{\text{reading_score}}{\text{passage_length}}
\]

`reading_score` is the number of words read correctly. Dividing by
`passage_length` puts surveys with different passage lengths on a common
scale: a score of 54 on a 60-word passage and a score of 81 on a 90-word
passage both give accuracy 0.9. The same formula is used for a second passage
as `readingB_accuracy` when that passage exists.

### Foundational reading skills

Foundational reading skills (`reading_skills`) is a binary indicator. It
is 1 only when *both* of the following hold; otherwise it is 0 among children
who have a reading score:

1.  All five comprehension questions are correct:
    \(\text{read_comp_score} = 5\).
2.  At least 90% of the passage words are correct. The 90% rule is applied to
    a whole number of words:

\[
\text{cutoff} = \lfloor 0.9 \times \text{passage_length} \rfloor
\]

\[
\text{reading_skills} =
\begin{cases}
1 & \text{if } \text{read_comp_score} = 5
  \text{ and } \text{reading_score} \ge \text{cutoff} \\
0 & \text{otherwise (when } \text{reading_score} \text{ is non-missing)}
\end{cases}
\]

So foundational skills is not “read well” alone: a child who reads every word
but misses a comprehension item fails, and a child who answers every
comprehension question but reads fewer than 90% of the words also fails. The
same rule applies to a second passage as `readingB_skills`.

In short: accuracy tracks continuous reading performance, while foundational reading skills marks children who both read at least $90\%$ of the passage and show full comprehension.


<p class="code-lang code-lang-stata">Stata</p>

```{stata eval=FALSE}
generate reading_accuracy = reading_score / passage_length
generate cutoff = int(0.9 * passage_length)
generate reading_skills = read_comp_score == 5 &                    ///
                          (reading_score >= cutoff & reading_score < .) if reading_score < .
```

<p class="code-lang code-lang-r">R</p>

```{r eval=FALSE}
d <- d %>% mutate(
  reading_accuracy = reading_score / passage_length,
  cutoff = trunc(0.9 * passage_length),
  reading_skills = if_else(
    present(reading_score),
    as.numeric(eq(read_comp_score, 5) & ge_obs(reading_score, cutoff)),
    NA_real_
  )
)
```

`ge_obs()` requires both sides to be non-missing, because Stata's
`reading_score >= cutoff` is false when the cutoff is missing.

## Collapse extra passages

Lesotho and Madagascar fielded three passages where every other survey used at
most two. Their third passage is folded into the `B` slots and the `C`
variables are dropped, so the appended file has a consistent two-passage
structure.

<p class="code-lang code-lang-stata">Stata</p>

```{stata eval=FALSE}
if "`iso'" == "LSO" | "`iso'" == "MDG" {
    replace readingB_score    = readingC_score    if readingC_score    < .
    replace readingB_accuracy = readingC_accuracy if readingC_accuracy < .
    replace readingB_skills   = readingC_skills   if readingC_skills   < .
    * ... the remaining C variables, then:
    drop read_compC* passageC_* readingC_* wordsC_* cutoffC practiceC*
}
```

<p class="code-lang code-lang-r">R</p>

```{r eval=FALSE}
if (iso %in% c("LSO", "MDG")) {
  for (i in seq_along(pairs)) {
    b  <- names(pairs)[i]
    cc <- unname(pairs[i])
    if (b %in% names(d) && cc %in% names(d)) {
      d[[b]] <- if_else(present(d[[cc]]), d[[cc]], d[[b]])
    }
  }
  d <- d %>% select(-matches("^(read_compC|passageC_|readingC_|wordsC_|practiceC)"))
}
```

## Classify the assessment outcome

Not every child has a reading score, and the reasons matter. A child may not
have been interviewed, may fall outside the target age range, may have had
consent withheld, may have been assessed in a language they do not use, may
have declined to read, or may have failed the practice. `reading_status`
records all of this on one variable.

Two flags feed into it. `lang_mismatch` marks children who consented but were
never offered a story, where the language of instruction (for those who
attended school) or the language spoken at home (for those who did not) is
recorded. `child_refuses_read` marks children who were offered a story and
declined, with country-specific additions for surveys that record refusal in a
different variable.

<p class="code-lang code-lang-stata">Stata</p>

```{stata eval=FALSE}
generate lang_mismatch = child_consent == 1 &                                  ///
    ((ever_attended == 1 & likestory == . & lang_school < .) |                  ///
     ((enrolled == 2 | ever_attended == 2) & likestory == . & lang_home < .))

generate child_refuses_read = likestory != 1 & likestory < .

egen max_reading_score    = rowmax(reading*_score)
egen max_reading_accuracy = rowmax(reading*_accuracy)
egen max_reading_skills   = rowmax(reading*_skills)

generate reading_status = 0 if interview_result > 1 & interview_result < .
replace  reading_status = 1 if age < 7 | (age > 14 & age < .)
replace  reading_status = 2 if consent != 1 & consent < . & reading_status >= .
replace  reading_status = 3 if child_consent != 1 & child_consent < . & reading_status >= .
replace  reading_status = 4 if lang_mismatch == 1 & reading_status >= .
replace  reading_status = 5 if child_refuses_read == 1
replace  reading_status = 6 if fail_practice == 1
replace  reading_status = 7 if max_reading_score < .
replace  reading_status = 8 if max_reading_skills == 1
```

<p class="code-lang code-lang-r">R</p>

```{r eval=FALSE}
d$max_reading_score    <- row_max(d, "^reading[A-Za-z]*_score$")
d$max_reading_accuracy <- row_max(d, "^reading[A-Za-z]*_accuracy$")
d$max_reading_skills   <- row_max(d, "^reading[A-Za-z]*_skills$")

d <- d %>% mutate(
  reading_status = if_else(present(interview_result) & interview_result > 1, 0, NA_real_),
  reading_status = if_else(present(age) & (age < 7 | age > 14), 1, reading_status),
  reading_status = if_else(ne_obs(consent, 1) & is.na(reading_status), 2, reading_status),
  reading_status = if_else(ne_obs(child_consent, 1) & is.na(reading_status), 3, reading_status),
  reading_status = if_else(eq(lang_mismatch, 1) & is.na(reading_status), 4, reading_status),
  reading_status = if_else(eq(child_refuses_read, 1), 5, reading_status),
  reading_status = if_else(eq(fail_practice, 1), 6, reading_status),
  reading_status = if_else(present(max_reading_score), 7, reading_status),
  reading_status = if_else(eq(max_reading_skills, 1), 8, reading_status)
)
```

The `max_reading_*` variables assign each child their best result across the
passages they attempted, so a child who read a second passage better than the
first is credited with the better outcome.

Read the ordering carefully, because it is not a plain priority list. Codes 2,
3 and 4 only apply where nothing has been assigned yet, so an earlier reason
wins. Codes 5 through 8 overwrite whatever came before, so the last applicable
rung wins. In practice this means attempting a passage (7) supersedes a
practice failure (6), and meeting the skills threshold (8) supersedes
everything.

| Code | Meaning |
|---|---|
| 0 | Not interviewed |
| 1 | Age under 7 or over 14 |
| 2 | Caregiver consent not given |
| 3 | Child consent not given |
| 4 | Assessment language does not match the child's language |
| 5 | Child does not want to read the story |
| 6 | Failed the practice sentence and questions |
| 7 | Attempted a reading passage |
| 8 | Read 90% of words and answered all comprehension questions correctly |

`reading_status` is missing for children who pass every gate but have no
recorded score. That is worth tabulating with `missing` when you first open the
file.

## Merge keys and variable selection

Each survey is uniquely identified by cluster, household and line number. The
scripts assert this before going further, then copy the three identifiers to
the names used by IPUMS-MICS extracts (`cluster`, `hhno`, `linech`), which is
what lets you merge these outcomes onto a harmonised IPUMS dataset later.

<p class="code-lang code-lang-stata">Stata</p>

```{stata eval=FALSE}
isid HH1 HH2 LN
clonevar cluster = HH1
clonevar hhno    = HH2
clonevar linech  = LN

generate country_iso3 = "`iso'"
generate year = `year'
```

<p class="code-lang code-lang-r">R</p>

```{r eval=FALSE}
if (anyDuplicated(d[c("HH1", "HH2", "LN")]) > 0) {
  stop("HH1 HH2 LN do not uniquely identify children in ", path)
}

d <- d %>% mutate(
  cluster = HH1, hhno = HH2, linech = LN,
  country_iso3 = iso, year = as.numeric(year)
)
```

The survey is then cut down to the analysis variables. Because questionnaires
differ, the Stata version expands each requested pattern and keeps only what
exists, rather than failing on an absent variable:

<p class="code-lang code-lang-stata">Stata</p>

```{stata eval=FALSE}
local tokeep ""
foreach v of local wanted {
    capture unab match : `v'
    if _rc == 0 local tokeep "`tokeep' `match'"
}
keep `tokeep'
```

<p class="code-lang code-lang-r">R</p>

```{r eval=FALSE}
d %>% select(
  any_of(c("country_iso3", "year", "cluster", "hhno", "linech",
           "HH1", "HH2", "LN", "reading_status", "age", "consent",
           "child_consent", "enrolled", "ever_attended", "lang_home",
           "lang_school", "lang_mismatch", "child_refuses_read",
           "fail_practice")),
  matches("^words"), matches("^cutoff"), matches("^max_reading_"),
  matches("^reading"), matches("^read_comp"), matches("^practice"),
  matches("^passage")
)
```

The kept set is wider than the derived outcomes alone. It retains the raw word
counts, the age and consent variables, and the intermediate flags, so you can
audit or re-derive `reading_status` without going back to the source files.

## Append, label, save

Each survey is written to a temporary file inside the loop, and the temporary
files are appended once at the end. Value and variable labels are applied to
the combined dataset, not to each survey, so the label definitions live in one
place.

<p class="code-lang code-lang-stata">Stata</p>

```{stata eval=FALSE}
use "`part1'", clear
forvalues i = 2/`n' {
    append using "`part`i''"
}

label define reading_status 0 "Not interviewed" 1 "Age <7 or >14" /// ...
label values reading_status reading_status

compress
save "`outpath'/`outfile'.dta", replace
```

<p class="code-lang code-lang-r">R</p>

```{r eval=FALSE}
harmonized <- bind_rows(parts) %>% apply_labels()

write_dta(harmonized, file.path(outdir, paste0(outfile, ".dta")))
```

Surveys that lack a second passage simply get missing values for the `B`
variables when appended.

## What the finished dataset contains

One row per child in the Foundational Learning Skills module, across all
supported surveys.

| Variable | Meaning |
|---|---|
| `country_iso3`, `year` | Survey identification |
| `cluster`, `hhno`, `linech` | Merge keys for IPUMS-MICS (copies of `HH1`, `HH2`, `LN`) |
| `reading_status` | Outcome of the reading assessment, coded 0–8 |
| `words_att`, `words_incorrect` | Raw word counts for passage 1 |
| `reading_score` | Words read correctly, passage 1 |
| `reading_accuracy` | `reading_score / passage_length` |
| `reading_skills` | 90% of words plus all comprehension correct |
| `read_comp_1`–`read_comp_5`, `read_comp_score` | Comprehension items and total |
| `practice_correct`, `practice_question1`, `practice_question2`, `practice_outcome` | Practice sentence and questions |
| `passage_language`, `passage_length` | Passage characteristics |
| `*B_*` equivalents | The same measures for the second passage |
| `max_reading_score`, `max_reading_accuracy`, `max_reading_skills` | Best result across passages |
| `age`, `consent`, `child_consent`, `enrolled`, `ever_attended`, `lang_home`, `lang_school` | Inputs to the classification |
| `lang_mismatch`, `child_refuses_read`, `fail_practice`, `cutoff`, `cutoffB` | Intermediate flags |



## Known limitations

Worth being aware of before you use the output:

- Passage lengths are hard-coded. They come from the national
  questionnaires, not from the data. If you add a survey, you must add its
  passage length, or accuracy and `reading_skills` cannot be computed.
- Tunisia has no passage language. Its length (72) is assigned, but the
  language coding scheme here has no Arabic code, so `passage_language` stays
  missing. Nothing downstream depends on it.
- Year differences within a country are not modelled. Tunisia appears twice
  (2018 and 2023) and both are given the same passage length.
- `reading_skills` is 0, not missing, when comprehension is missing but a
  reading score exists. This follows the original definition; treat such
  cases carefully if they are numerous in your surveys.
- `reading_status` can be missing for children who pass every gate without
  a recorded score.

## Checks after running

<p class="code-lang code-lang-stata">Stata</p>

```{stata eval=FALSE}
use "data/mics6_reading_harmonized.dta", clear

tabulate country_iso3 year
tabulate reading_status, missing
summarize reading_accuracy
```

<p class="code-lang code-lang-r">R</p>

```{r eval=FALSE}
library(tidyverse)
library(haven)

d <- read_dta("data/mics6_reading_harmonized.dta")

count(d, country_iso3, year)
count(d, reading_status)                       # includes NA
summary(d$reading_accuracy)
d %>% filter(reading_status == 8) %>% nrow()
```

Accuracy should sit between 0 and 1. Values above 1 point to a passage length
that is too short for the survey in question, which is the first thing to check
when adding a country.

