## Creating a survey design for one country survey {#c05-single-survey}

For one MICS6 survey the selected-child design is:

``` text
weight   = fsweight
PSU      = PSU
stratum  = stratum
```


### Check the design variables

Restrict to one survey first. Then count missing weights, PSUs, and strata, and how many PSUs sit in each stratum.

::::: {.code-tabs}
:::: {.code-panel data-lang="Stata"}
```stata
use "Data/AFLEARN Harmonised Data/mics6-fs-harmonised-v2.0.dta", clear
keep if country_iso3 == "GHA"

misstable summarize fsweight PSU stratum
* If PSU is absent, use HH1 in its place.
svydescribe
```
::::
:::: {.code-panel data-lang="R"}
```{r eval=FALSE}
library(haven)
library(dplyr)

d = read_dta("Data/AFLEARN Harmonised Data/mics6-fs-harmonised-v2.0.dta") |>
  filter(country_iso3 == "GHA")    # subset to Ghana survey

d |>
  summarise(
    n = n(),
    missing_weight = sum(is.na(fsweight)),
    missing_psu = sum(is.na(PSU)),
    missing_stratum = sum(is.na(stratum)),
    n_psu = n_distinct(PSU),
    n_strata = n_distinct(stratum)
  )
```
::::
:::::


### Declare the design

`nest = TRUE` tells R that PSU codes can repeat across strata.

::::: {.code-tabs}
:::: {.code-panel data-lang="Stata"}
```stata
svyset PSU [pw=fsweight], strata(stratum)
svydescribe
```
::::
:::: {.code-panel data-lang="R"}
```{r eval=FALSE}
library(srvyr)
des = d |>
  as_survey_design(
      ids = ~PSU,
   strata = ~stratum,
  weights = ~fsweight,
     data = d,
     nest = TRUE
  )
  
```
::::
:::::


### Lonely PSUs

Variance within a stratum needs more than one PSU. A small domain can leave a stratum with one contributing PSU. Inspect that before changing options. In R, a documented default when lonely domains occur is:

```{r eval=FALSE}
options(
  survey.lonely.psu = "adjust",
  survey.adjust.domain.lonely = TRUE
)
```

In Stata, read singleton-unit messages rather than ignoring them.
