## Setting up one survey {#c05-single-survey}

For one MICS6 survey the selected-child design is:

``` text
weight   = fsweight
PSU      = PSU
stratum  = stratum
```

If `PSU` and `stratum` are missing from the AFLEARN file, merge them from `Data/UNICEF/COUNTRY_SURVEY/fs.sav` ([Adding variables from UNICEF files](#c03-merge)), or use `HH1` as the cluster and nest it inside the survey.

If you do not yet have the analysis file, start from [Getting UNICEF MICS files](#c01-unicef).

### Check the design variables

Restrict to one survey first. Then count missing weights, PSUs, and strata, and how many PSUs sit in each stratum.

::::: {.code-tabs}
:::: {.code-panel data-lang="Stata"}
```stata
use "Data/AFLEARN Harmonised Data/mics6-fs-harmonised-v2.0.dta", clear
keep if country_iso3 == "GHA"

misstable summarize fsweight PSU stratum
* If PSU is absent, use HH1 in its place.
svydescribe
```
::::
:::: {.code-panel data-lang="R"}
```{r eval=FALSE}
library(haven)
library(dplyr)

d <- read_dta("Data/AFLEARN Harmonised Data/mics6-fs-harmonised-v2.0.dta") %>%
  filter(country_iso3 == "GHA")

d %>%
  summarise(
    n = n(),
    missing_weight = sum(is.na(fsweight)),
    missing_psu = sum(is.na(PSU)),
    missing_stratum = sum(is.na(stratum)),
    n_psu = n_distinct(PSU),
    n_strata = n_distinct(stratum)
  )
```
::::
:::::

Prefer the released `PSU` to `HH1`. They often agree; `PSU` is the sampling-error variable. IPUMS names them `PSU` and `STRATUM`. Check the names in your extract.

### Declare the design

`nest = TRUE` tells R that PSU codes can repeat across strata.

::::: {.code-tabs}
:::: {.code-panel data-lang="Stata"}
```stata
svyset PSU [pw=fsweight], strata(stratum)
svydescribe
```
::::
:::: {.code-panel data-lang="R"}
```{r eval=FALSE}
library(survey)

des <- svydesign(
  ids = ~PSU,
  strata = ~stratum,
  weights = ~fsweight,
  data = d,
  nest = TRUE
)
```
::::
:::::

`srvyr::as_survey_design()` wraps the same specification if you want dplyr verbs.

### Estimate a mean

The mean of a 0/1 variable is the estimated proportion equal to 1.

::::: {.code-tabs}
:::: {.code-panel data-lang="Stata"}
```stata
svy: mean reading_skills
svy: mean numeracy_score
```
::::
:::: {.code-panel data-lang="R"}
```{r eval=FALSE}
svymean(~reading_skills, design = des, na.rm = TRUE)
svymean(~numeracy_score, design = des, na.rm = TRUE)
```
::::
:::::

### Restrict to ages 7–14 after declaring the design

Define the design on the full selected-child sample, then restrict the design object. Dropping rows first discards information the variance estimator needs.

::::: {.code-tabs}
:::: {.code-panel data-lang="Stata"}
```stata
svy, subpop(if age >= 7 & age <= 14): mean reading_skills
```
::::
:::: {.code-panel data-lang="R"}
```{r eval=FALSE}
des_7_14 <- subset(des, age >= 7 & age <= 14)
svymean(~reading_skills, design = des_7_14, na.rm = TRUE)
```
::::
:::::

### Lonely PSUs

Variance within a stratum needs more than one PSU. A small domain can leave a stratum with one contributing PSU. Inspect that before changing options. In R, a documented default when lonely domains occur is:

```{r eval=FALSE}
options(
  survey.lonely.psu = "adjust",
  survey.adjust.domain.lonely = TRUE
)
```

In Stata, read singleton-unit messages rather than ignoring them.
