# Harmonising MICS6 FS background themes {#c06-harmonize-fs}

Cross-country analysis of schooling, labour, discipline, functioning, and
numeracy needs one child-level file with stable names. Official MICS6
`fs.sav` files use different layouts and country codes, so they do not
pool cleanly on their own.

[Harmonising MICS6 reading outcomes](#c05-harmonize) already builds reading
scores from the Foundational Learning module. This chapter covers everything
else from the children aged 5–17 (FS) questionnaire: geography, parent
education, schooling background, labour, parental involvement, functioning,
discipline, and numeracy setup. Reading item batteries stay in the reading
file; merge the two files later on shared keys.

One R script does the work: `harmonize-mics-fs-v1.0.R`. You do not need the
full guide repository. Follow [Download the MICS Data](#c01-download) to
build `data/MICS_Datasets/`, place the script beside that folder, and run it.

```{r eval=FALSE}
source("data/harmonize-mics-fs-v1.0.R")
```

The script writes:

| Output | Role |
|---|---|
| `data/mics6_fs_harmonized.rds` | Analysis file (R) |
| `data/mics6_fs_harmonized.dta` | Analysis file (Stata) |
| `data/mics6_fs_variable_crosswalk.xlsx` | Source map, exclusions, level maps, notes |

The sections below explain each theme so you can adapt the logic rather than
treat the script as a black box.

## What you need

1. Complete [Download the MICS Data](#c01-download) so each survey folder
   under `data/MICS_Datasets/` holds `fs.sav`. Parent education and school
   type also need `hl.sav` in the same folder.
2. Obtain `harmonize-mics-fs-v1.0.R` (for example from the AFLEARN GitHub
   assets for this guide) and save it as `data/harmonize-mics-fs-v1.0.R`.
3. Set the R working directory to the folder that **contains** `data/`
   (not to `data/` itself).
4. Have R with packages the script loads through `pacman`: `tidyverse`,
   `haven`, `labelled`, `openxlsx`.

Your tree should look like this:

``` text
<your-project>/
└── data/
    ├── MICS_Datasets/
    │   └── BEN_2021_MICS6_v01_M/
    │       ├── fs.sav
    │       └── hl.sav
    └── harmonize-mics-fs-v1.0.R
```

The script recognises these ISO3 codes (Tunisia appears twice, 2018 and 2023):

``` text
BEN CAF COD COM GHA GMB GNB LSO MDG MWI NGA SLE STP SWZ TCD TGO TUN ZWE
```

Any other survey folder is skipped with a message.

## Run the script

From the project folder that contains `data/`:

```{r eval=FALSE}
source("data/harmonize-mics-fs-v1.0.R")
```

The run prints one block per survey (children kept, exclusion counts). On a
full set of supported surveys the pooled file has about **166,980** children
and **135** variables. Exact counts depend on which surveys you prepared.

If Excel has `mics6_fs_variable_crosswalk.xlsx` open, Windows may lock the
file. Close it before re-running, or use the `_updated.xlsx` / `.tmp.xlsx`
copy the script leaves when it cannot replace the locked workbook.

## The pipeline at a glance

``` text
for each survey folder
  1.  read ISO3 and year from the folder name
  2.  read fs.sav once (and hl.sav for parent education / school type)
  3.  apply themes on the child row:
        HH  geography and household size
        WT  FS sample weights (fsweight, fshweight)
        INT FS interview date and start/end times
        BG  mother/father education + school type
        CB  schooling background (+ level_h)
        CL  child labour
        PR  parental involvement
        FCF child functioning
        FCD child discipline
        FL  numeracy items + FL setup (not reading outcomes)
  4.  keep merge keys and theme columns
append all surveys
apply value and variable labels
write .rds, .dta, and the Excel crosswalk
```

Identifiers `HH1`, `HH2`, and `LN` must uniquely identify children in each
`fs.sav`. The script also stores copies as `cluster`, `hhno`, and `linech`
for merges with IPUMS-MICS or the reading file.

## Geography and household size

**Depth:** rename and labels only.

| Harmonized | Source | Meaning |
|---|---|---|
| `urban` | `HH6` | Urban (1) / rural (2) |
| `region` | `HH7` | First administrative area (**country codes**) |
| `n_children_5_17` | `HH52` | Number of children aged 5–17 in the household |

`region` is not a cross-country taxonomy. Interpret codes within
`country_iso3`. `n_children_5_17` is missing in Tunisia 2018. Extra HH
fields (`HH3`, `HH4`, `HH6A`, `HH7A`, …) are listed on the Excel `excluded`
sheet.

## Sample weights

**Depth:** keep MICS names and labels only (no wealth indices).

| Harmonized | Source | Meaning |
|---|---|---|
| `fsweight` | `fsweight` | Children 5–17 sample weight |
| `fshweight` | `fshweight` | Children 5–17 household sample weight |

Both fields are present in every supported survey in this build. Wealth
scores (`wscore`, `windex*`) are not included.

## Interview date and time

**Depth:** rename and English labels only (no derived duration or date).

| Harmonized | Source | Meaning |
|---|---|---|
| `interview_day` | `FS7D` | Day of FS interview |
| `interview_month` | `FS7M` | Month of FS interview |
| `interview_year` | `FS7Y` | Year of FS interview |
| `interview_start_hour` | `FS8H` | Start hour |
| `interview_start_min` | `FS8M` | Start minutes |
| `interview_end_hour` | `FS11H` | End hour |
| `interview_end_min` | `FS11M` | End minutes |

All seven fields are on every supported `fs.sav` in this build. Household
interview dates (`HH5D`/`HH5M`/`HH5Y`) are not included.

## Parent education and school type

**Depth:** raw parent education codes plus a common `*_edu_h` map (same
0–5 / 8 / 9 scheme as CB `*_level_h`). School type is rename only.

| Harmonized | Source | Meaning |
|---|---|---|
| `mother_edu` | FS `melevel` | Mother's education (country constructed codes) |
| `mother_edu_h` | mapped from `mother_edu` | Harmonised level |
| `father_edu` | HL `felevel` | Father's education (merged on `HH1`/`HH2`/`LN`=`HL3`) |
| `father_edu_h` | mapped from `father_edu` | Harmonised level |
| `school_type` | HL `ED11` | School ownership/type for the FS child |

`felevel` does not appear on `fs.sav`; the script always reads it from
`hl.sav`. `school_type` is missing in Gambia and Togo.

Most surveys code school type as public (1), religious (2), private (3),
community (4), other (6). **COD** uses a different network taxonomy—keep
raw codes and read the Excel `notes` sheet. Tunisia codes “other” as 4,
not 6.

Parent `*_edu_h` maps live in the script (`parent_level_map_for`) and on
the Excel `level_map` sheet (`source_var` = `melevel` or `felevel`). Code
9 covers missing, “no information,” and parent not in the household. Malawi
raw code 5 means vocational for mothers and father-not-in-HH for fathers.

## Child's Background (schooling)

**Depth:** rename plus harmonised education levels `highest_level_h`,
`current_level_h`, `previous_level_h`.

| Harmonized | Source |
|---|---|
| `birth_month`, `birth_year` | `CB2M`, `CB2Y` |
| `age` | `CB3` |
| `ever_attended` | `CB4` |
| `highest_level`, `highest_grade`, `highest_completed` | `CB5A`, `CB5B`, `CB6` |
| `enrolled`, `current_level`, `current_grade` | `CB7`, `CB8A`, `CB8B` |
| `attended_previous`, `previous_level`, `previous_grade` | `CB9`, `CB10A`, `CB10B` |

Raw level and grade variables keep country codes. Grade value labels are
generic English (`Grade/year 1`, …, plus DK / no response /
inconsistent); they do not claim cross-country grade equivalence. Raw
`*_level`, parent education, `school_type`, `region`, and language fields
have **no pooled value labels**—codes stay numeric; use `*_level_h` /
`*_edu_h` and the Excel `level_map` / `notes` sheets for meaning.

**Warning:** `current_grade` is the year **within** `current_level_h`, not
the child's place in the school system. Primary year 1 and secondary year 1
both carry the code `1`. See
[Why current_grade is not the child's grade](#c07-current-grade).

The `*_level_h` variables use:

| Code | Label |
|---:|---|
| 0 | Early childhood / pre-primary (ECE) |
| 1 | Primary |
| 2 | Lower secondary |
| 3 | Upper secondary |
| 4 | Vocational / technical |
| 5 | Higher / tertiary |
| 8 | Don't know |
| 9 | No response / other special |

Country maps sit in `level_map_for()` and on the Excel `level_map` sheet.
Defaults in v1.0: nested tech/voc inside secondary → 3; standalone
vocational tracks → 4; Zimbabwe vocational/tertiary subtypes collapse to
4/5; Tunisia adult education / literacy → 9.

## Child Labour

**Depth:** rename and labels only (no ILO/MICS derived labour indicators).

Core block: economic work (`work_*`), hours, hazards, water/firewood, and
household chores. Tunisia 2018 has no CL module—the script creates those
columns as missing. Country extras (herding, night work, fishing, …) go to
the `excluded` sheet.

## Parental Involvement

**Depth:** rename and labels only.

Core items cover books, homework help, school governing body / PTA
meetings, report cards, school visits, school closure reason slots, teacher
absence, and contact with officials.

**Warning:** `school_closed_a` / `b` / `c` (PR12A–C) are **not**
cross-country comparable as coded. Slot meanings differ by survey (for
example Nigeria puts COVID-19 in slot A). Read the Excel `notes` sheet
before pooling those three columns. Use `school_closed_other` (PR12X) for
“any other reason.”

## Child Functioning

**Depth:** rename and labels only (no Washington Group “any difficulty”
composites).

Core aids, seeing/hearing, walking, self-care, communication, cognition,
psychosocial, and affect items (`uses_glasses` … `sad_depressed_freq`).
Walking distance wording says “yards” in some surveys and “meters” in
others; codes are treated as comparable. Child Discipline (`FCD*`) is a
separate theme, not folded into FCF exclusions.

## Child Discipline

**Depth:** rename and labels only (no violent-discipline composites).

| Harmonized | Source |
|---|---|
| `disc_took_privileges` … `disc_beat_hard` | `FCD2A`–`FCD2K` |
| `phys_punish_needed` | `FCD5` (SLE: `FCD3`) |

Sierra Leone has no `FCD5`; the attitude item sits in `FCD3` and maps to
`phys_punish_needed`. Filters `FCD3`/`FCD4` (where not consumed) and
country extras `FCD2L`–`N` are excluded.

## Foundational Learning (numeracy only)

**Depth:** rename and labels only (no numeracy skill scores).

| Block | Harmonized examples | Sources |
|---|---|---|
| Setup | `fl_consent`, `child_consent`, `lang_home`, `lang_school`, `fl_child_result` | `FL1`, `FL3`, `FL7`, `FL9`/`FL9A`+`FL9B`, `FL28` |
| Numeracy | `number_id_*`, `number_compare_*`, `number_add_*`, `number_pattern_*` | `FL23*`, `FL24*`, `FL25*`, `FL27*` |

`lang_school` prefers `FL9`; otherwise `FL9A` with `FL9B` fill (for example
Chad). `fl_child_result` is missing in Sierra Leone.

All reading FL items (word lists, practice, comprehension, B/C passages)
are **excluded** here. Use `mics6_reading_harmonized` and merge on
`country_iso3`, `year`, `HH1`, `HH2`, `LN`.

## Documentation the script writes

`data/mics6_fs_variable_crosswalk.xlsx` has four sheets:

| Sheet | Content |
|---|---|
| `crosswalk` | Theme, harmonized name, label, notes, and per-survey source |
| `excluded` | Source variables left out, with reasons |
| `level_map` | Raw → `level_h` / parent `*_edu_h` maps |
| `notes` | Theme caveats (PR12, COD school type, SLE discipline, …) |

Design history for maintainers also lives under
`docs/fs-harmonization-plans/` in the published guide repository. You do
not need that folder to run the script.

## What the harmonized dataset contains

One row per child in the FS module across supported surveys you prepared.

| Block | Examples |
|---|---|
| Keys | `country_iso3`, `year`, `cluster`, `hhno`, `linech`, `HH1`, `HH2`, `LN` |
| HH | `urban`, `region`, `n_children_5_17` |
| WT | `fsweight`, `fshweight` |
| INT | `interview_day`, `interview_month`, `interview_year`, start/end hour and minutes |
| BG | `mother_edu`, `mother_edu_h`, `father_edu`, `father_edu_h`, `school_type` |
| CB | schooling fields + `*_level_h` |
| CL | work, hazards, chores, hours |
| PR | books, school engagement, closure slots |
| FCF | functioning items |
| FCD | discipline methods + attitude |
| FL | numeracy items + FL setup |

## Known limitations

- `region`, raw education levels/grades, and `school_type` keep country
  codes; do not pool them as a single taxonomy without remapping. Pooled
  SPSS value labels are cleared for those fields (and for languages /
  line identifiers) so French or mixed dictionaries from the first survey
  do not appear in the shared file.
- `current_grade`, `highest_grade`, and `previous_grade` restart within
  each education level. Pair them with `*_level_h` before treating them as
  years of schooling — see
  [Why current_grade is not the child's grade](#c07-current-grade).
- The script does not derive child-labour, violent-discipline, functional
  difficulty, or foundational numeracy skill indicators.
- Reading outcomes are not duplicated; merge the reading file when you need
  them.
- Coverage of father education and school type depends on `hl.sav` and on
  how often those fields are filled on the child's household-list row.

## Checks after running

```{r eval=FALSE}
library(tidyverse)
library(haven)

d <- read_rds("data/mics6_fs_harmonized.rds")
# or: d <- read_dta("data/mics6_fs_harmonized.dta")

count(d, country_iso3, year)
ncol(d)

# Urban / rural
count(d, urban)

# Harmonised mother education
count(d, mother_edu_h)

# SLE attitude item should not be all missing
d %>% filter(country_iso3 == "SLE") %>%
  summarise(n = n(), phys_nonmiss = sum(!is.na(phys_punish_needed)))

# GMB and TGO: school_type should be missing
d %>% filter(country_iso3 %in% c("GMB", "TGO")) %>%
  summarise(school_type_nonmiss = sum(!is.na(school_type)))
```

## Next steps

With `data/mics6_fs_harmonized.dta` (or `.rds`) in place:

- Read
  [Why current_grade is not the child's grade](#c07-current-grade) before
  using `current_grade` as a year of schooling.
- Build or refresh reading outcomes with
  [Harmonising MICS6 reading outcomes](#c05-harmonize) and merge on
  `country_iso3`, `year`, `HH1`/`HH2`/`LN` (or `cluster`/`hhno`/`linech`).
- Follow [R Tutorial](#c03-r) or [Stata Tutorial](#c04-stata) for analysis.

The same keys also support merges onto an IPUMS-MICS extract for additional
household background variables.
