# Why current_grade is not the child's grade {#c07-current-grade}

`current_grade` looks like a grade. It is not. In the harmonised FS file it is
the year **within** the education level in `current_level_h`. Primary year 1
and secondary year 1 both carry the code `1`. Treat that code as the child's
place in the school system and you pool children who are nowhere near the same
year of schooling.

For ages 5–17, enrolment by year of schooling should rise through primary and
thin through secondary. The raw `current_grade` distribution does something
else: it piles every cycle onto the same 1, 2, 3 ticks and leaves a sawtooth.

```{r setup-grade, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 9,
  fig.height = 5,
  fig.align = "center"
)

fs_path <- "data/mics6_fs_harmonized.rds"
has_fs <- file.exists(fs_path)
theme_path <- "R/aflearn_theme.R"
```

```{r packages-grade, eval=has_fs}
library(tidyverse)
library(srvyr)
library(haven)

source(theme_path)
```

```{r load-grade, eval=has_fs}
fs <- readRDS(fs_path)

an <- fs %>%
  mutate(
    survey = paste(country_iso3, year),
    enrolled_n = as.integer(haven::zap_labels(enrolled)),
    grade = as.integer(haven::zap_labels(current_grade)),
    level = as.integer(haven::zap_labels(current_level_h)),
    HH1 = as.integer(haven::zap_labels(HH1)),
    fsweight = as.numeric(haven::zap_labels(fsweight))
  ) %>%
  filter(
    enrolled_n == 1L,
    !is.na(grade), grade %in% 1:20,
    !is.na(level), level %in% 0:5,
    !is.na(fsweight), fsweight > 0,
    !is.na(HH1)
  )

# Illustrative plots: primary + lower/upper secondary only
an_main <- an %>% filter(level %in% 1:3)

level_lab <- c(
  "1" = "Primary",
  "2" = "Lower secondary",
  "3" = "Upper secondary"
)

design_pooled <- function(df) {
  df %>%
    as_survey_design(
      ids = HH1,
      strata = survey,
      weights = fsweight,
      nest = TRUE
    )
}

design_one <- function(df) {
  df %>%
    as_survey_design(
      ids = HH1,
      weights = fsweight,
      nest = TRUE
    )
}
```

```{r no-data-note, eval=!has_fs, echo=FALSE, results='asis'}
cat(
  "> Harmonised FS file not found at `data/mics6_fs_harmonized.rds`. ",
  "Figures below are skipped. Run ",
  "[Harmonising MICS6 FS background themes](#c06-harmonize-fs) first.\n"
)
```

## What you see

Pooled across the supported MICS6 surveys, the weighted share of currently
enrolled children by `current_grade` spikes at 1 and falls steeply after 3.
That is not a primary-to-secondary pyramid. It is every cycle restarting its
count on the same axis.

```{r plot-sawtooth, eval=has_fs, fig.cap="Weighted share of currently enrolled children by current_grade. Grade 1 is highlighted: that bar mixes primary, lower secondary, and upper secondary."}
share_grade <- design_pooled(an_main %>% filter(grade %in% 1:9)) %>%
  group_by(grade) %>%
  summarise(
    p = survey_mean(vartype = NULL),
    n = unweighted(n()),
    .groups = "drop"
  )

n_saw <- sum(share_grade$n)

share_grade %>%
  mutate(
    fill = if_else(grade == 1L, aflearn_amber, aflearn_neutral[["context"]])
  ) %>%
  ggplot(aes(x = factor(grade), y = p, fill = fill)) +
  geom_col(width = 0.72, colour = NA) +
  scale_fill_identity() +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0.08))
  ) +
  labs(
    title = "current_grade piles every cycle onto 1, 2, 3",
    subtitle = paste0(
      "Weighted share of enrolled children · primary and secondary · ",
      "unweighted n = ", format(n_saw, big.mark = ",")
    ),
    x = "current_grade",
    y = NULL,
    caption = "Source: harmonised MICS6 FS. Design: clusters = HH1, strata = survey, weights = fsweight."
  ) +
  theme_aflearn() +
  theme(axis.text.x = aflearn_axis_category())
```

## Why the bumps

`current_grade` is MICS `CB8B`: grade or year **within** the level in `CB8A`.
The harmonised sibling of that level is `current_level_h`. Value labels on the
grade field even say "Grade/year 1".

Stack the same shares by level and the sawtooth breaks apart. Most of the mass
at 1 is primary. A large second slice is lower secondary year 1. A third is
upper secondary year 1. One code, three places in the system.

| `current_level_h` | `current_grade` | Meaning |
|---|---:|---|
| Primary (1) | 1 | Year 1 of primary — often the first year of school |
| Lower secondary (2) | 1 | Year 1 of lower secondary — not grade 1 of school |
| Upper secondary (3) | 1 | Year 1 of upper secondary |

```{r plot-stacked, eval=has_fs, fig.cap="Same current_grade axis, stacked by current_level_h. The spike at 1 is three cycles, not one grade."}
share_stack <- design_pooled(an_main %>% filter(grade %in% 1:9)) %>%
  # interact() so cell shares sum to 1 overall (not within each grade)
  group_by(interact(grade, level)) %>%
  summarise(
    p = survey_mean(vartype = NULL),
    n = unweighted(n()),
    .groups = "drop"
  ) %>%
  mutate(
    level_f = factor(
      level_lab[as.character(level)],
      levels = c("Primary", "Lower secondary", "Upper secondary")
    )
  )

cols_level <- aflearn_pal_categorical(3)
names(cols_level) <- c("Primary", "Lower secondary", "Upper secondary")

share_stack %>%
  ggplot(aes(x = factor(grade), y = p, fill = level_f)) +
  geom_col(width = 0.72, colour = NA) +
  scale_fill_manual(values = cols_level, name = NULL) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0.08))
  ) +
  labs(
    title = "Grade 1 is primary — and secondary — and upper secondary",
    subtitle = "Weighted cell shares · stacked · currently enrolled children",
    x = "current_grade",
    y = NULL,
    caption = "Source: harmonised MICS6 FS. ECE, vocational, and tertiary omitted from the plot."
  ) +
  theme_aflearn() +
  theme(
    legend.position = "top",
    axis.text.x = aflearn_axis_category()
  )
```

Numbering is not even uniform across surveys. Most restart at 1 in each cycle.
Tunisia already codes lower secondary as 7–9. Primary runs to 8 in Malawi, 7 in
Lesotho, Eswatini, and Zimbabwe, 5 in Madagascar, and 9 in Guinea-Bissau. The
pooled bumps hide those differences as well.

## The fix

Always read `current_grade` with `current_level_h`. To place a child on one
ladder of years of schooling, add a country-specific offset for each level that
restarts its count:

\[
\text{admin\_grade} = \text{offset}(\text{country}, \text{level}) + \text{current\_grade}
\]

Where the survey already continues the count (Tunisia lower secondary 7–9), use
`current_grade` as-is. Leave vocational and tertiary off the general ladder:
those tracks do not map cleanly onto primary–secondary years.

## After the fix: Ghana

Ghana's structure is clean for a demonstration: six years of primary, then
lower and upper secondary that each restart at 1 (a 6+3+3 pattern in this
file). Build an administrative year as primary = `current_grade`, lower
secondary = `6 + current_grade`, upper secondary = `9 + current_grade`.

The left panel still shows the sawtooth. The right panel shows enrolment by
constructed year: a rise through primary and a thin secondary tail.

```{r ghana-admin, eval=has_fs}
gha <- an_main %>%
  filter(country_iso3 == "GHA") %>%
  mutate(
    admin_grade = case_when(
      level == 1L ~ grade,
      level == 2L ~ 6L + grade,
      level == 3L ~ 9L + grade,
      TRUE ~ NA_integer_
    )
  ) %>%
  filter(!is.na(admin_grade))

gha_des <- design_one(gha)

share_raw <- gha_des %>%
  filter(grade %in% 1:6) %>%
  group_by(grade) %>%
  summarise(
    p = survey_mean(vartype = NULL),
    n = unweighted(n()),
    .groups = "drop"
  ) %>%
  transmute(x = grade, p, n, panel = "current_grade (raw)")

share_admin <- gha_des %>%
  group_by(admin_grade) %>%
  summarise(
    p = survey_mean(vartype = NULL),
    n = unweighted(n()),
    .groups = "drop"
  ) %>%
  transmute(x = admin_grade, p, n, panel = "admin_grade (6+3+3)")

gha_plot <- bind_rows(share_raw, share_admin) %>%
  mutate(
    panel = factor(
      panel,
      levels = c("current_grade (raw)", "admin_grade (6+3+3)")
    )
  )
```

```{r plot-ghana, eval=has_fs, fig.height=5.5, fig.cap="Ghana 2017: raw current_grade still saw-toothed; admin_grade restores a primary-to-secondary enrolment shape."}
gha_plot %>%
  ggplot(aes(x = factor(x), y = p)) +
  geom_col(width = 0.72, fill = aflearn_electric_blue, colour = NA) +
  facet_wrap(~panel, scales = "free_x", ncol = 2) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0.08))
  ) +
  labs(
    title = "Same Ghana children, two grade axes",
    subtitle = "Weighted shares · currently enrolled · primary and secondary",
    x = NULL,
    y = NULL,
    caption = "Source: harmonised MICS6 FS, Ghana 2017. Design: clusters = HH1, weights = fsweight."
  ) +
  theme_aflearn() +
  theme(axis.text.x = aflearn_axis_category())
```

### Offsets for the other surveys

These lengths are **empirical in this file**, not a new official taxonomy. Check
them before you borrow an offset.

| Pattern | Surveys (examples) | How to build `admin_grade` |
|---|---|---|
| Primary 6, secondary restarts at 1 | Benin, Ghana, Nigeria, many others | Primary = grade; add 6 for lower secondary; add 9 for upper secondary (adjust if lower secondary has a different length) |
| Primary longer than 6 | Malawi (to 8); Lesotho, Eswatini, Zimbabwe (to 7) | Use that survey's max primary grade as the offset into secondary |
| Primary shorter | Madagascar (to 5) | Offset secondary by 5 |
| Count continues | Tunisia lower secondary 7–9 | Use `current_grade` for that level; do not add a primary offset |
| No separate lower-secondary code | Guinea-Bissau, Lesotho, Eswatini, São Tomé | `current_level_h` jumps from primary to upper secondary / other; invent no fake lower-secondary years |

## Same trap elsewhere

`highest_grade` and `previous_grade` restart within `highest_level_h` and
`previous_level_h` the same way. Pair each grade with its level. Do not force
vocational or tertiary years onto the primary–secondary ladder.

## Recipe

Restrict to currently enrolled children with a usable grade. Compute shares with
`srvyr`. Build the administrative year with country-specific offsets.

<p class="code-lang code-lang-r">R</p>

```{r recipe-r, eval=FALSE}
library(tidyverse)
library(srvyr)
library(haven)

fs <- readRDS("data/mics6_fs_harmonized.rds")

gha <- fs %>%
  mutate(
    enrolled_n = as.integer(haven::zap_labels(enrolled)),
    grade = as.integer(haven::zap_labels(current_grade)),
    level = as.integer(haven::zap_labels(current_level_h)),
    HH1 = as.integer(haven::zap_labels(HH1)),
    fsweight = as.numeric(haven::zap_labels(fsweight))
  ) %>%
  filter(
    country_iso3 == "GHA",
    enrolled_n == 1L,
    grade %in% 1:20,
    level %in% 1:3,
    !is.na(fsweight), fsweight > 0
  ) %>%
  mutate(
    admin_grade = case_when(
      level == 1L ~ grade,
      level == 2L ~ 6L + grade,
      level == 3L ~ 9L + grade
    )
  )

gha %>%
  as_survey_design(ids = HH1, weights = fsweight, nest = TRUE) %>%
  group_by(admin_grade) %>%
  summarise(
    p = survey_mean(vartype = NULL),
    n = unweighted(n())
  )
```

<p class="code-lang code-lang-stata">Stata</p>

```{stata eval=FALSE}
use "data/mics6_fs_harmonized.dta", clear
keep if country_iso3 == "GHA"
keep if enrolled == 1
keep if inrange(current_grade, 1, 20)
keep if inlist(current_level_h, 1, 2, 3)

generate admin_grade = .
replace admin_grade = current_grade if current_level_h == 1
replace admin_grade = 6 + current_grade if current_level_h == 2
replace admin_grade = 9 + current_grade if current_level_h == 3

svyset HH1 [pw = fsweight]
svy: tabulate admin_grade
```

## Next steps

- Build or refresh the FS file with
  [Harmonising MICS6 FS background themes](#c06-harmonize-fs).
- For reading outcomes, see
  [Harmonising MICS6 reading outcomes](#c05-harmonize).
- Merge on `country_iso3`, `year`, and `HH1` / `HH2` / `LN` (or
  `cluster` / `hhno` / `linech`).
